---
title: "Binomial test"
author:
  name: "Anne Stempel" 
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
abstract: > 
  A binomial distribution is a statistical distribution which can be found in an experiment with only two possible outcomes (`0` and `1`, also referred to as `failure` and `success`, respectively) and which are independent of each other in case of repetition.
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
---

## Recommended reading

For linguists:

> Gries [-@gries_statistics_2021: 39--45]

General:

> Baguley [-@baguleySeriousStatsGuide2012: Chapters 2.3.1, 4.5.1, 4.8.4]


## Linguistic Example of a Binomial Distribution

To illustrate this, consider a binomial distribution in linguistics,
such as gender bias in pronoun use. The research question is as follows:

> Do people tend to exhibit gender bias in pronoun use (e.g. *he* or
> *she*) when referring to professionals in leadership roles?

As a result, one can either expect to see pronouns used in a way that
aligns with traditional gender stereotypes, or notice efforts to
counteract such biases by adopting more inclusive and neutral language.
Nevertheless, the pronouns that are being analysed can only be female or
male and thus, they represent the required structure for a binomial
distribution (*he* or *she*), which is binary data.

A test that is built on the assumptions of the binomial distribution is
called the **Binomial Test**. The Binomial Test calculates the likelihood of
observing a particular pronoun (e.g., *he*) being used to refer to
professionals in leadership roles, under the assumption of a given
expected proportion, and assesses whether the observed frequency differs
significantly from the expected one.

::: callout-tip
## Is this approach suitable for my research?

If you are working with corpus-based frequency data that involves binary
categories (yes or no, male or female, VO or OV, British or
US-American,...) and if you have a specific probability of success which
is specified in your hypotheses and does not change from one trial to
the next, then this approach is working for you and will lead to
successful results.
:::

## Understanding the Binomial Distribution (Cumulative Mass Function (CMF) Version)

The binomial distribution is used to calculate the probability of
observing up to and including $x$ successes in $n$ trials, given a fixed
probability of success $P$. This calculation is conducted using the
**Cumulative Mass Function** (CMF), which is a running total of
probabilities. The CMF is a valuable tool for calculating probabilities
for ranges, such as "what is the probability of 0 to 3 successes?" as
opposed to the probability of a specific number of successes (see
@eq-binom). [^binomial_test-1]

[^binomial_test-1]: For more information on the binomial test using a
    specific number of successes, please refer to The Probability Mass
    Function PMF.

$$ P(x;n,P) = \sum_{i=1}^{|x|}\bigg[\binom{n}{i} p^i (1-P)^{n-i}\bigg]
$$ {#eq-binom}

-   Key elements:
    -   $x$: The maximum number of successes that is of interest (e.g.,
        up to 3 successes).
    -   $n$: The total number of trials (e.g., total observations or
        experiments).
    -   $P$: The probability of success in each trial (e.g., the chance
        of using *he* when describing a professional).

::: callout-tip
## Linguistic example

Imagine studying gender bias in pronoun use. If $P = 0.5$ (i.e., the
probability of obtaining the masculine pronoun), and you observe up to 3
uses of *he* out of 10 trials, the CMF calculates the probability of
getting 0, 1, 2, or 3 successes (usage of *he*) combined.
:::

## Using the Binomial Test (CMF-Version) in Linguistic Research

The CMF is especially useful when testing hypotheses about ranges of
outcomes.

Steps for a Binomial Test Using the CMF:

1. Hypotheses

-   Null hypothesis ($H_0$): The observed data matches the expected
    cumulative probability. Therefore: The observed frequency of the
    pronouns *he* and *she* referring to professionals in leadership
    roles corresponds with the expected proportion of 50% each,
    indicating that both pronouns are being used equally.

-   Alternative hypothesis ($H_1$): The observed data deviates from
    the expected cumulative probability. Therefore: The observed
    frequency of the pronouns *he* and *she* referring to professionals
    in leadership roles deviates from the expected proportion of 50%
    each, indicating a gender bias.

2. Performing the test in R by using the cumulative probabilities

-   Example: To calculate the probability of observing up to 3 successes
    in 10 trials with P=0.5 each, the following code should be used:

```         
pbinom(3,10,0.5)
```

3\. Interpreting the results:

-   A p-value smaller than 0.05 indicates a significant deviation from
    the expected probability

-   The test helps to determine whether the observed data can be used to
    reject the H0.

-   Checking the confidence interval (CI) provides the range of likely
    values for the true success probability

::: callout-tip
## The Probability Mass Function (PMF)

This function is used to calculate the probability of observing exactly
x successes in n trials, given a fixed probability of success P.

$$ 
\begin{align}
f(x; n, P) & = \binom{n}{x} P^x (1-P)^{n-x} \\
\end{align}
$$

To calculate the PMF binomial distributions in R, the following code
should be used:

```{r}         
dbinom(x = 3, size = 10, prob = 0.5)
```

:::

## One-Sided vs. Two-Sided Tests

The choice between one-sided and two-sided tests depends on the research
question.

-   One-Sided Test: Used only for analysing deviations in one direction.

    -   Example: Is a participant's performance better than chance?

-   Two-Sided Test: Used for analysing deviations in both directions.

    -   Example: Is the frequency of male pronouns different from the
        frequency of female pronouns? (The result can either be that the
        frequency of male pronouns is higher than the frequency of
        female pronouns or that they are less frequent than female
        pronouns.)

Two-Sided Binomial Test

```{r}         
binom.test(x = 15, n = 20, p = 0.5, alternative = "two.sided")
```

One-Sided Binomial Test

```{r}         
binom.test(x = 15, n = 20, p = 0.5, alternative = "greater")
```

::: callout-tip
## Please note

The function `binom.test()` in R only uses the CMF function.
:::

## Comparing Two Groups: Proportion Tests

For the comparison of proportions between two groups, the prop.test()
function can be used.

Example: An experiment is being conducted, in which participants have to
determine whether a word they are being shown is an actual lemma of the
English language or not.

-   Group A achieved 24 correct answers out of 25 trials

-   Group B obtained 19 correct answers out of 25 trials.

-   To see, whether the proportions are significantly different, the
    following code can be used.

```{r}         
prop.test(c(24,19), c(25,24))
```

## Testing Event Counts Over Time: The Poisson Test

For analysing event counts over time, such as pauses in speech, the
**Poisson distribution** with the `poisson.test()` function can be useful.

**Example**: An experiment is conducted in which speakers are asked to
talk for a predetermined time. and it is being researched how often
pauses occur in speech.

-   Participant A pauses 18 times in 5 minutes.

-   The researcher expects 2 pauses per minute.

-   To see whether the observed pause rate differs from the expected
    pause rate, the following test is highly convenient.

```{r}        
poisson.test(18, T=5, r=2)
```
