---
title: "Libraries"
author:
  name: "Vladimir Buskin" 
  orcid: "0009-0005-5824-1012"
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
    
editor: visual
---

## Recommended reading

> @winter_statistics_2020: Chapter 1.13

## Working with packages in R

Packages expand the basic functionality of R by providing numerous
quality-of-life improvements that not only considerably simplify common
data wrangling tasks but which also provide frameworks for
state-of-the-art methods for statistical analysis and natural language
processing (NLP), among many other things.

### Installation {#sec-libraries}

::: {.callout-note collapse="false" title="How do I install a library?"}
Navigate to `Packages` \> `Install` and verify that the pop-up window
says `Install from: Repository (CRAN)`. You can now type in the name of
the package you would like to install under `Packages`.

[Video tutorial on YouTube](https://www.youtube.com/watch?v=u1r5XTqrCTQ)
:::

This reader will use functions from a variety of R packages. Please
install the following ones:

| Package      | Purpose                                           | Session                                                                             |
|------------------------|------------------------|------------------------|
| `readxl`     | Importing Microsoft Excel files                   | [Importing/exporting](Importing_exporting.qmd)                                      |
| `writexl`    | Exporting Microsoft Excel files                   | [Importing/exporting](Importing_exporting.qmd)                                      |
| `quanteda`   | Analysis of text data                             | [Concordancing](Concordancing.qmd)                                                  |
| `lattice`    | Data visualisation                                | [Concordancing](Concordancing.qmd)                                                  |
| `tidyverse`  | Framework for data manipulation and visualisation | [Categorical data](Categorical_data.qmd), [Continuous data](Summary_statistics.qmd) |
| `crosstable` | Creating contingency tables                       | [Categorical data](Categorical_data.qmd)                                            |
| `flextable`  | Exporting contingency tables                      | [Categorical data](Categorical_data.qmd)                                            |
| `confintr`   | Effect size measure for categorical data          | [Chi-squared test](Chi_square_test.qmd)                                             |

### Loading packages

Once the installation has been completed, you can proceed to load the
libraries using the code below. You can ignore the warning messages.

```{r, echo = TRUE}

library(readxl)
library(writexl)
library(quanteda)
library(lattice)
library(tidyverse)
library(crosstable)
library(flextable)
library(confintr)
```

::: {.callout-note collapse="false" title="Activating libraries"}
Whenever you start a new R session (i.e., open RStudio), your libraries
and their respective functions will be inactive. To re-activate a
library, either use the `library()` function or simply select it in the
`Packages` tab.
:::

It is good practice to only activate those packages that are necessary
for your analysis. While it won't be a problem for the small set of
packages as shown here, loading dozens of packages increases the risk of
obtaining "homonymous" functions which have the same name but perform
different operations. In this case, it might be helpful to
"disambiguate" them by directly indicating which package a function is
from:

```{r, echo = TRUE, output = FALSE, eval = FALSE}
readxl::read_xlsx(...)
```

### Citing R and R packages

Whenever we draw on ideas other than our own, we give credit to the
respective source by citing it appropriately. The same applies to R,
RStudio as well as all the packages we rely on throughout our analyses.

For R, an up-to-date citation can be generated as follows:

```{r}
citation()
```

To cite a specific package, simply supply the package name as an
argument.

```{r}
citation("quanteda")
```

```{r, include = FALSE, eval = FALSE}
sessionInfo()
```

```{r, include = FALSE, eval = FALSE}
library(report)

cite_packages()
```
