---
title: "Libraries"
author: Vladimir Buskin
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
    
editor: visual
---

## Installing and loading packages

Packages expand the basic functionality of R by providing numerous
quality-of-life improvements that not only considerably simplify common
data wrangling tasks but which also provide frameworks for
state-of-the-art methods for statistical analysis and natural language
processing (NLP).

In order to install a package, you navigate to `Packages` \> `Install`
and verify that the pop-up window says
`Install from: Repository (CRAN)`. You can now type in the name of the
package you would like to install under `Packages`.

**Task**: Install the `tidyverse` and `quanteda` packages!

Once the installation has been completed, you can proceed to load the
libraries using the code below. You can ignore the warning messages.

```{r, echo = TRUE}

library(tidyverse)
library(quanteda)
```

```{r, echo = FALSE}

library(kableExtra)

```

Remember that you will have to reload these libraries whenever you start
a new R session (i.e., open RStudio).
