---
title: "Linear regression"
author:
  name: "Vladimir Buskin" 
  orcid: "0009-0005-5824-1012"
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
format:
  html:
    self-contained: true
    logo: logo.png
    footer: "Regression"
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 4
    scrollable: true
editor: visual
bibliography: R.bib
---

## Recommended reading

For linguists:

> Levshina [-@levshina_how_2015: Chapter 7]
>
> Winter [-@winter_statistics_2020: Chapter 4]

General:

> Heumann et al. [-@heumann_introduction_2022: Chapter 11]
>
> James et al. [-@james_introduction_2021: Chapter 3]

## Preparation {.smaller}

```{r, echo = TRUE, output = TRUE}
# Load libraries
library(readxl) # for reading in Excel data
library(tidyverse) # data manipulation and visualisation framework
library(broom) # converting models to data frames
library(sjPlot) # exporting regression tables
library(effects) # plot marginal effects
library(ggeffects) # generating predictions
library(car) # model diagnostics

# Load data
ELP <- read_xlsx("ELP.xlsx")

# Inspect data structure
str(ELP)
```

## Introduction

Consider the distribution of the variables `RT` (reaction times) and
`Freq` from the `ELP` (English Lexicon Project) dataset
[@balotaEnglishLexiconProject2007].

We will apply a $\log$-transformation to both variables in order to even
out the differences between extremely high and extremely low frequency
counts [@winter_statistics_2020: 90-94].

::: panel-tabset
### Log-transformed

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Code"

# Log-transformed
ggplot(ELP, aes(x = log(RT))) +
  geom_histogram() +
  geom_vline(xintercept = mean(log(ELP$RT)), color = "steelblue") +
  geom_vline(xintercept = mean(log(ELP$RT)), color = "red") +
  theme_minimal() +
  labs(
    x = "Reaction time (log)"
  ) +
   annotate("text", x = log(mean(ELP$RT)), y = 0, 
           label = "mean", 
           vjust = 1.5, hjust = -0.2, color = "steelblue", parse = TRUE) +
  annotate("text", x = log(mean(ELP$RT)) + -0.2, y = .7, 
           label = "median", 
           vjust = 1.5, hjust = -0.2, color = "red", parse = TRUE)


```

### Untransformed

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Code"

# Skewed
ggplot(ELP, aes(x = RT)) +
  geom_histogram() +
  geom_vline(xintercept = mean(ELP$RT), color = "steelblue") +
  geom_vline(xintercept = median(ELP$RT), color = "red") +
  theme_minimal() +
  labs(
    x = "Reaction time"
  ) +
annotate("text", x = mean(ELP$RT) + 10, y = 0, 
           label = "mean", 
           vjust = 1.5, hjust = -0.2, color = "steelblue", parse = TRUE) +
  annotate("text", x = median(ELP$RT) -175, y = .7, 
           label = "median", 
           vjust = 1.5, hjust = -0.2, color = "red", parse = TRUE)

```
:::

We are particularly interested in the relationship between reaction
times `RT` and the (log-)frequency `Freq` of a lexical stimulus. What
kind of pattern does the scatter plot below suggest?

```{r, echo = FALSE, output = TRUE}
ggplot(ELP, aes(x = log(Freq), y = log(RT))) +
  geom_point() +
  theme_minimal() +
  labs(
    x = "Log-transformed word frequency",
    y = "Log-transformed reaction time"
  )
```

::: {.callout-caution title="Some open questions" collapse="false"}
-   Can word frequency help us **explain variation** in reaction times?

-   If it can, then how could we characterise the **effect** of word
    frequency? In other words, does it increase or decrease reaction
    times?

-   What reaction times should we expect for **new observations**?
:::

### A simple statistical model

`RT` is the *response* or *target* that we wish to explain. We
generically refer to the response as $Y$.

`Freq` is the *feature*, *input*, or *predictor*, which we will call
$X$.

We can thus summarise our preliminary and fairly general statistical
model as

$$Y = f(X) + \epsilon.
$$ {#eq-model}

The term $f(X)$ describes the contribution of $X$ to the explanation of
$Y$. Since no model can explain everything perfectly, we expect there to
be some degree of error $\epsilon$.

## Linear regression

-   Linear regression is a simple approach to supervised machine
    learning where the response variable is known.[^linear_regression-1]

-   It assumes that the dependence of $Y$ on $X$ is **linear**, i.e.,
    their relationship is a straight line.

-   This approach is suitable for **numerical response variables**. The
    predictors, however, can be either continuous or discrete.

[^linear_regression-1]: If the response variable is unknown or
    irrelevant, we speak of **unsupervised machine learning**.
    Unsupervised models are mostly concerned with finding patterns in
    high-dimensional datasets with dozens or even hundreds of variables.

Although it may seem overly simplistic, linear regression is **extremely
useful** both conceptually and practically.

### Model with a single predictor $X$ {.smaller}

The simple linear model has the general form

$$ Y = \beta_0 + \beta_1X + \epsilon. 
$$ {#eq-linear}

-   The model parameters (or coefficients) $\beta_0$ and $\beta_1$
    specify the functional relationship $f$ between $Y$ and $X$.

-   The first parameter $\beta_0$ determines the **intercept** of the
    regression line, and $\beta_1$ indicates the **slope**.

-   Once again, $\epsilon$ captures the model error, which is equivalent
    to the sum of all distances of the data points from the regression
    line.

```{r, echo = F}
# Create sample data
set.seed(123)
n <- 50
x <- runif(n, -5, 10)
y <- 2 + 0.5 * x + rnorm(n, sd = 1)
data <- data.frame(x = x, y = y)

# Fit linear regression
model <- lm(y ~ x, data = data)
intercept <- coef(model)[1]
slope <- coef(model)[2]

# Find a point with a large residual
large_residual <- data %>%
  mutate(residual = abs(y - predict(model))) %>%
  arrange(desc(residual)) %>%
  dplyr::slice(1)


# Assuming the data and model are already created as in the previous example

# Choose points for the slope triangle
x1 <- 2
x2 <- 6
y1 <- intercept + slope * x1
y2 <- intercept + slope * x2

ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "grey20") +
  geom_segment(aes(xend = x, yend = predict(model)), color = "grey10", alpha = 0.5) +
  geom_vline(xintercept = 0, col = "grey10", linetype = "dashed") +
  
  # Intercept annotation
  annotate("text", x = 0.7, y = intercept + 1.75, label = expression(beta[0]), 
           hjust = 0, vjust = 2, color = "purple") +
  annotate("curve", x = 0.5, y = intercept + 1, xend = 0, yend = intercept,
           arrow = arrow(length = unit(0.3, "cm")), curvature = 0.3, color = "purple") +
  
  # Slope annotation with triangle
  geom_segment(x = x1, xend = x2, y = y1, yend = y1, color = "darkred", linetype = "dashed") +
  geom_segment(x = x2, xend = x2, y = y1, yend = y2, color = "darkred", linetype = "dashed") +
  annotate("text", x = (x1 + x2)/2, y = y1 - 0.5, label = expression(Delta * x), color = "darkred") +
  annotate("text", x = x2 + 0.3, y = (y1 + y2)/2, label = expression(Delta * y), color = "darkred", angle = 90) +
  annotate("text", x = x2 + 1, y = y2 - 1.5, label = expression(beta[1] == frac(Delta * y, Delta * x)), 
           color = "darkred", hjust = 0) +
  
  # Improved Residuals annotation
  annotate("text", x = large_residual$x, y = large_residual$y, 
           label = expression(epsilon), 
           hjust = -1, vjust = -5, color = "red") +
  geom_segment(aes(x = large_residual$x, xend = large_residual$x, 
                   y = large_residual$y, yend = predict(model, newdata = large_residual)),
               color = "red", linetype = "dashed") +
  
  theme_minimal() +
  labs(title = "Elements of the Linear Regression Model",
       x = "X", y = "Y")


```

Applying the model formula to our dataset, we get the following updated
regression equation:

$$ \text{Reaction time} = \beta_0 + \beta_1\text{Frequency} + \text{Model Error}. $$

But how do we find the exact values of the intercept and the slope? In
short: We can't! We are dealing with population parameters and can,
therefore, only provide an approximation of the true relationship
between `RT` and `Freq`.

To reflect the tentative nature of the model coefficients, we use the
**hat** symbol \^ (e.g., $\hat{\beta_0}$) to indicate **estimations**
rather than true values. The estimation procedure requires **training
data**, based on which the algorithm "learns" the relationship between
$Y$ and $X$ (hence the term "Machine Learning").

::: {.callout-tip title="How exactly do you estimate the model parameters?" collapse="true"}
The most common way of estimating parameters for linear models is the
**Least Squares** approach. In essence, the parameters are chosen such
that the residual sum of squares, i.e., the sum of the differences
between observed and predicted values, is as low as possible. In other
words, we are trying to minimise the distances between the data points
and the regression line.

It can be computed using the equivalence in @eq-lse-slope.

$$ \hat{\beta}_1 = \frac{\sum_{i=1}^n(x_i - \bar{x})(y_i- \bar{y})}{\sum_{i=1}^n(x_i- \bar{x})^2}.
$$ {#eq-lse-slope}

We can then obtain the intercept via @eq-lse-intercept.

$$
\hat{\beta_0}= \bar{y}- \hat{\beta}_1\bar{x}
$$ {#eq-lse-intercept}
:::

Once we've fitted the model, we can then predict reaction times if we
know the frequency of a lexical stimulus:

$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1x,
$$ {#eq-preds}

where $\hat{y}$ indicates a prediction of $Y$ on the basis of the
predictor values $X = x$.

### Application in R

In R, we can fit a linear model with the `lm()` function.

```{r, echo = TRUE, output = FALSE}
# Fit linear model
rt.lm1 = lm(log(RT) ~ log(Freq), data = ELP)

# View model data
summary(rt.lm1)
```

```{r, echo = TRUE, output = TRUE}
tab_model(rt.lm1, show.se = TRUE, show.aic = TRUE, show.dev = TRUE, show.fstat = TRUE, digits = 3)
```

The model statistics comprise the following elements:

::: {.callout-tip title="Call" collapse="true"}
i.e., the model formula.
:::

::: {.callout-tip title="Residuals" collapse="true"}
These indicate the difference between the observed values in the data
set and the values predicted by the model (= the fitted values). These
correspond to the error term $\epsilon$. The lower the residuals, the
better the model describes the data.

```{r, echo = TRUE, output = TRUE}

# Show fitted values (= predictions) for the first six observations
head(rt.lm1$fitted.values)

# Show deviation of the fitted values from the observed values
head(rt.lm1$residuals)
```
:::

::: {.callout-tip title="Coefficients" collapse="true"}
The regression coefficients correspond to $\hat{\beta}_0$ ("Intercept")
and $\hat{\beta}_1$ ("log(Freq)"), respectively. The model shows that
for a one-unit increase in log-frequency the log-reaction time decreases
by approx. 0.05.

```{r, echo = TRUE, output = TRUE, message = FALSE, warning = FALSE}

# Convert coefficients to a tibble (= tidyverse-style data frame)
tidy_model <- tidy(rt.lm1)

tidy_model

```
:::

::: {.callout-tip title="$p$-values and $t$-statistic" collapse="true"}
$p$**-values and** $t$**-statistic**: Given the null hypothesis $H_0$
that there is no correlation between `log(RT)` and `log(Freq)` (i.e.,
$H_0: \beta_1 = 0$), a $p$-value lower than 0.05 indicates that
$\beta_1$ considerably deviates from 0, thus providing evidence for the
alternative hypothesis $H_1: \beta_1 \ne 0$. Since $p < 0.001$, we can
reject $H_0$.

The $p$-value itself crucially depends on the
$t$-statistic[^linear_regression-2], which measures "the number of
standard deviations that $\hat{\beta_1}$ is away from 0"
[@james_introduction_2021: 67]. The standard error (SE) reflects how
much an estimated coefficient differs on average from the true values of
$\beta_0$ and $\beta_1$. They can be used to compute the 95% confidence
interval
$$[\hat{\beta}_1 - 2 \cdot SE(\hat{β}_1), \hat{\beta}_1 + 2 \cdot SE(\hat{\beta}_1)].
$$ {#eq-ci}

The true value of the parameter $\beta_1$ lies within the specified
range 95% of the time.

```{r, echo = TRUE, output = TRUE}

# Compute confidence intervals for intercept and log(Freq)
tidy_model_ci <- tidy(rt.lm1, conf.int = TRUE)

tidy_model_ci

```

The estimated parameter for `log(Freq)`, which is -0.049, thus has the
95% confidence interval \[-0.053, -0.044\].
:::

[^linear_regression-2]: If the response variable is unknown or
    irrelevant, we speak of **unsupervised machine learning**.
    Unsupervised models are mostly concerned with finding patterns in
    high-dimensional datasets with dozens or even hundreds of variables.

::: {.callout-tip title="**Residual standard error** (RSE)" collapse="true"}
This is an estimation of the average deviation of the predictions from
the observed values.

$$RSE = \sqrt{\frac{1}{n-2}\sum_{i=1}^n (y_i - \hat{y_i}})^2
$$ {#eq-rse}
:::

::: {.callout-tip title="$R^2$" collapse="true"}
The $R^2$ score is important for assessing model fit because it
"measures the proportion of variability in $Y$ that can be explained
using $X$" [@james_introduction_2021: 70], varying between 0 and 1.

$$R^2 = 1-\frac{TSS}{RSS} = 1-\frac{\sum_{i=1}^n (y_i - \hat{y_i})^2}{\sum_{i=1}^n (y_i - \bar{y_i})^2}
$$ {#eq-r2}
:::

::: {.callout-tip title="$F$-statistic" collapse="true"}
It is used to measure the association between the dependent variable and
the independent variable(s). Generally speaking, values greater than 1
indicate a possible correlation. A sufficiently low $p$-value suggests
that the null hypothesis $H_0: \beta_1 = 0$ can be rejected. The $F$
statistic is computed as shown below [cf.
@agrestiFoundationsStatisticsData2022: 232] and follows an
$F$-distribution with two different $df$ values.

$$
F = \frac{(TSS - SSE) / p}{SSE / [n - (p + 1)]}
$$ {#eq-f}
:::

### Multiple linear regression

In multiple linear regression, more than one predictor variable is taken
into account. For instance, modelling `log(RT)` as a function of
`log(Freq)`, `POS` and `Length` requires a more complex model of the
form

$$ Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_pX_p + \epsilon.
$$ {#eq-multreg}

Predictions are then obtained via the formula

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1x_1 + \hat{\beta}_2x_2 + ... + \hat{\beta}_px_p.
$$ {#eq-multreg-preds}

### Application in R

In R, a multiple regression model is fitted as in the code example
below:

```{r, echo = TRUE, output = FALSE, warning = FALSE, message = FALSE}
# Fit multiple regression model
rt.lm2 <- lm(log(RT) ~ log(Freq) + POS + Length, data = ELP)

# View model statistics
summary(rt.lm2)
```

```{r, echo = TRUE, output = TRUE, warning = FALSE, message = FALSE}
tab_model(rt.lm2, show.se = TRUE, show.aic = TRUE, show.dev = TRUE, show.fstat = TRUE, digits = 3)
```

## Visualising regression models

-   Plot coefficient estimates:

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Code"

# Tidy the model output
tidy_model <- tidy(rt.lm2, conf.int = TRUE)

# Remove intercept
tidy_model <- tidy_model %>% filter(term != "(Intercept)")

# Create the coefficient plot
ggplot(tidy_model, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "steelblue") +
  theme_minimal() +
  labs(
    x = "Coefficient Estimate",
    y = "Predictor",
    title = "Coefficient Estimates with Confidence Intervals"
  )

```

-   Plot predictions:

```{r, echo = TRUE, output = TRUE, warning = FALSE, message = FALSE}
#| code-fold: true
#| code-summary: "Code"

plot(ggeffect(rt.lm2, "Freq"), residuals = TRUE) + geom_line(col = "steelblue") + labs(subtitle = "Untransformed frequencies", y = "log(RT)")

plot(ggeffect(rt.lm2, "Freq [log]"), residuals = TRUE) + geom_line(col = "steelblue") + labs(subtitle = "Log-transformed frequencies", y = "log(RT)")

plot(ggeffect(rt.lm2, "POS"), residuals = TRUE) + geom_line(col = "steelblue") + labs(y = "log(RT)")

plot(ggeffect(rt.lm2, "Length"), residuals = TRUE) + geom_line(col = "steelblue") + labs(y = "log(RT)")
```

## Model assumptions and diagnostics

As a parametric method, linear regression makes numerous assumptions
about the training data. It is, therefore, essential to run further
tests to rule out possible violations. Among other things, the model
assumptions include:

-   A **linear relationship** between the response and the quantitative
    predictors: The residuals should not display a clear pattern. For
    this reason, it is recommended to use component residual plots
    (e.g., `crPlot()` from the `car` library) for the visual
    identification of potentially non-linear trends.

```{r}
#| code-fold: true
#| code-summary: "Code"

# pink line = main tendency vs. blue line = slope coefficients;
# some minor non-linearity can be observed

crPlot(rt.lm2, var = "log(Freq)") 
crPlot(rt.lm2, var = "POS")
crPlot(rt.lm2, var = "Length") # potentially problematic

```

-   **No heteroscedasticity** (i.e, non-constant variance of error
    terms): Visually, a violation of this assumption becomes apparent if
    the residuals form a funnel-like shape. It is also possible to
    conduct a non-constant variance test `ncvTest()`: If it returns
    $p$-values \< 0.05, it suggests non-constant variance.

```{r}
#| code-fold: true
#| code-summary: "Code"

plot(rt.lm2, which = 1)

ncvTest(rt.lm2) # significant, meaning that errors do not vary constantly

```

-   **No multicollinearity**: Predictors should not be correlated with
    each other. In the model data, correlated variables have unusually
    high standard errors, thereby decreasing the explanatory power of
    both the coefficients and the model as a whole. Another diagnostic
    measure are variance inflation factors (VIF-scores); predictors with
    VIF scores \> 5 are potentially collinear. They can be computed
    using the `vif()` function.

```{r}
#| code-fold: true
#| code-summary: "Code"

vif(rt.lm2) # vif < 5 indicates that predictors are not correlated
```

-   **Normally distributed residuals**: The residuals should follow the
    normal distribution and be centered around 0:

$$
\epsilon \sim N(0, \sigma^2)
$$ {#eq-erorr-distrib}

Usually, a visual inspection using `qqnorm()` is sufficient, but the
Shapiro-Wilke test `shapiro.test()` can also be run on the model
residuals. Note that a $p$-value below 0.05 provides evidence for
non-normality.

```{r}
#| code-fold: true
#| code-summary: "Code"

plot(rt.lm2, which = 2)

shapiro.test(residuals(rt.lm2)) # residuals are not normally distributed because p < 0.05
```

::: callout-important
Beside the points mentioned above, it is always recommend to examine the
model with regard to

-   **outliers** that might skew the regression estimates,

```{r}
#| code-fold: true
#| code-summary: "Code"

influencePlot(rt.lm2, id.method = "identify")
```

-   **interactions**, i.e., combined effects of predictors, and

```{r}
#| code-fold: true
#| code-summary: "Code"

rt.lm.int <- lm(log(RT) ~ log(Freq) + POS + Length + log(Freq):Length, data = ELP)

summary(rt.lm.int)

# ANOVA (analysis of variance)

## Compare interaction model with main effects model

anova(rt.lm.int, rt.lm2) # interaction term improves the model
```

-   **overfitting**, which results in poor model performance outside the
    training data.

```{r}
#| code-fold: true
#| code-summary: "Code"

library("rms")

# Refit the model with ols(), which is equivalent to lm()
ols.rt <- ols(log(RT) ~ log(Freq) + POS + Length, data = ELP, x = TRUE, y = TRUE)

# Cross-validate
ols.val <- validate(ols.rt, bw = TRUE, B = 200) # Perform 200 random resampling iterations (= bootstrapping); compare model performance on training vs. test (= new) data. The slope optimism should be below 0.05 to rule out overfitting.


ols.val[,1:5] # The model does not overfit.
```
:::
