---
title: "Mixed-effects regression"
author: Vladimir Buskin
format:
  html:
    self-contained: true
    logo: logo.png
    footer: "Regression"
    theme: Reference
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 4
    scrollable: true
editor: visual
bibliography: R.bib
---

## Recommended reading

For linguists:

> @paquot_mixed-effects_2020

General:

> @gelmanDataAnalysisUsing2007

## Preparation

```{r}
# Load libraries
library(tidyverse)
library(broom)
library(broom.mixed)
library(tidyr)
library(lme4) # for linear mixed-effects models
library(sjPlot)
library(ggeffects)
library(ggpubr)

# Load data
varmorph <- read.csv("varmorph_data.csv", header = TRUE)

# Reduce data
varmorph %>%
  select(rt, target, prime_type, subj_id) %>%
  filter(prime_type != "filler") %>% 
  drop_na() -> varmorph2

# Overview
glimpse(varmorph2)
```

## Multilevel models {#sec-mer}

At their core, mixed-effects models "are extensions of regression in
which data are structured in groups and coefficients can vary by group"
[@gelmanDataAnalysisUsing2007: 237]. Typical grouping structures found
in linguistic data include speakers, regions, or lexical stimuli for
which **multiple observations** are attested. Normally, such structures
would violate the assumption of independence, but can be controlled for
by capturing group-wise tendencies.

For illustration, a simple example of a hierarchical dataset is
presented in @fig-multi-simple. If one were to, for instance, measure
test scores for every student, it may be of interest how their
performance varies not only from student to student but also from school
to school. After all, the students are **nested** within their schools.

::: {#fig-multi-simple}
```{mermaid}
%%| fig-width: 4
%%| fig-height: 4
graph LR
    A1[School 1] --> B11(Student 1)
    A1[School 1] --> B12(Student 2)
    A1[School 1] --> B13(Student 3)
    
    A2[School 2] --> B21(Student 4)
    A2[School 2] --> B22(Student 5)
    A2[School 2] --> B23(Student 6)
    
```
:::

::: callout-caution
## Task

Read the following (partial) description of the experiments conducted by
Ciaccio & Veríssimo on the morphological processing of complex lexical
items [-@ciaccioInvestigatingVariabilityMorphological2022]:

> Sixty-nine intermediate to advanced non-native speakers of English (54
> women; 15 men) took part in the experiment in exchange for payment or
> course credits. \[...\] The experiment included 102 English
> monomorphemic verbs used as targets (e.g., *print*). These were
> preceded by their -*ed* past-tense form (e.g., *printed*) as the
> inflected prime, their -*er* nominalization (e.g., *printer*) as the
> derived prime, or by an unrelated prime. Unrelated primes were
> dissimilar in form and meaning from their corresponding targets; half
> of them were -*ed* inflected forms and half of them were -*er* derived
> words. [@ciaccioInvestigatingVariabilityMorphological2022: 2267]

Inspect `varmorph2` and characterise its multilevel structure.
:::

### Types of mixed-effects models

Variance across groups can be captured by **varying-intercept** and/or
**varying-slope** models. These varying coefficients also known as
**random effects** (cf. @gelmanDataAnalysisUsing2007: 245). In the model
equation, the intercept $\alpha$ and/or the slope $\beta$ is
additionally indexed for the grouping factor. Let $J$ denote the number
of groups for $j = 1, ..., J$.

::: callout-note
## Varying-intercept model

We allow group-wise variation in the intercept by replacing $\alpha$
with $\alpha_{j}$ to indicate the intercept for the $j$-th group. It is
defined as a random variable and follows the normal distribution. For
instance, each participant in the aforementioned psycholinguistic would
receive its own intercept rather than a global one for all participants.

$$
Y = \alpha_{j} + \beta_1X_{1} + \beta_2X_{2} + ... + \epsilon \qquad \alpha_{j} \sim N(\mu_{\alpha}, \sigma_{\alpha}^2) 
$$ {#eq-rnd-intercept}
:::

::: callout-note
## Varying-slope model

We will allow group-wise variation in the slope coefficients by
replacing them with $\beta_{ij}$ to indicate the slope for the $j$-th
group. The slope now functions as a random variable and is normally
distributed. In the psycholinguistic study, each participant would be
assigned its own slope coefficient.

$$
Y = \alpha + \beta_{1j}X_{1} + \beta_{2j}X_{2} + ... + \epsilon \qquad \beta_{j} \sim N(\mu_{\beta}, \sigma_{\beta}^2) 
$$ {#eq-rnd-slope}
:::

### Example

Assume we are predicting test performance by `School`. Using simulated
data, the following series of plots plots illustrate ...

(1) ... `School` as a fixed effect,

(2) ... random intercepts for each `School`,

(3) ... random slopes for each `School`, and

(4) ... random intercepts and slopes for each `School`.

```{r, echo = F}
# Set seed for reproducibility
set.seed(123)

# Generate sample data
generate_data <- function(n_schools = 5, n_students = 20) {
  schools <- data.frame(
    school = paste0("School ", LETTERS[1:n_schools]),
    intercept = rnorm(n_schools, mean = 60, sd = 5),
    slope = rnorm(n_schools, mean = 3, sd = 0.5)
  )
  
  data <- expand.grid(
    school = schools$school,
    student = 1:n_students
  ) %>%
    mutate(
      study_time = runif(n(), 0, 10),
      school_effect = schools$intercept[match(school, schools$school)] +
        schools$slope[match(school, schools$school)] * study_time,
      test_score = school_effect + rnorm(n(), mean = 0, sd = 5)
    )
  
  return(list(data = data, schools = schools))
}

# Generate data
data_list <- generate_data()
data <- data_list$data
schools <- data_list$schools

# Fit mixed-effects model
model <- lmer(test_score ~ study_time + (1 + study_time | school), data = data)

# Extract fixed effects
fixed_effects <- fixef(model)

# Function to plot
plot_mixed_effects <- function(data, model, show_random_intercepts = TRUE, show_random_slopes = TRUE) {
  # Extract random effects
  ranef_df <- as.data.frame(ranef(model)$school)
  ranef_df$school <- rownames(ranef_df)
  
  # Prepare data for school-specific lines
  school_lines <- schools %>%
    left_join(ranef_df, by = "school") %>%
    mutate(
      intercept = if (show_random_intercepts) intercept + `(Intercept)` else fixed_effects[1],
      slope = if (show_random_slopes) slope + study_time else fixed_effects[2]
    )
  
  # Create plot
  p <- ggplot(data, aes(x = study_time, y = test_score, color = school)) +
    geom_point(alpha = 0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", col = "grey10") +
    geom_abline(data = school_lines, 
                aes(intercept = intercept, slope = slope, color = school),
                linewidth = 1) +
    geom_abline(intercept = fixed_effects[1], slope = fixed_effects[2], 
                color = "black", linewidth = 1.5) +
    scale_color_brewer(palette = "Set1") +
    labs(x = "Study Time (hours)", y = "Test Score",
         title = "Mixed Effects Regression Visualisation",
         subtitle = paste("Random Intercepts:", show_random_intercepts, 
                          "| Random Slopes:", show_random_slopes)) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  return(p)
}

# Generate plots
p1 <- plot_mixed_effects(data, model, TRUE, TRUE)
p2 <- plot_mixed_effects(data, model, TRUE, FALSE)
p3 <- plot_mixed_effects(data, model, FALSE, TRUE)
p4 <- plot_mixed_effects(data, model, FALSE, FALSE)
```

::: {.callout-note title="Plot: (1) Fixed-effects model" collapse="true"}
```{r, echo = F}
print(p4)
```
:::

::: {.callout-note title="Plot: (2) Varying-intercept model" collapse="true"}
```{r, echo = F}
print(p2)
```
:::

::: {.callout-note title="Plot: (3) Varying-slope model" collapse="true"}
```{r, echo = F}
print(p3)
```
:::

::: {.callout-note title="Plot: (4) Varying-intercept and varying-slope model" collapse="true"}
```{r, echo = F}
print(p1)
```
:::

### Linear mixed-effects models

### Application in R

#### Varying-intercept model

```{r, output = F}
# Varying intercept model

# Define reference level for "prime_type"
varmorph2$prime_type <- factor(varmorph2$prime_type, levels = c("unrelated", "derived", "inflected"))

# Fit mixed-effects models
varmorph.me <- lmer(rt ~ prime_type + # fixed effect
                      (1 | subj_id) + # let intercept vary by subject
                      (1 | target), # # let intercept vary by target word
                      data = varmorph2)

# Summarise results
summary(varmorph.me)
```

```{r}
tab_model(varmorph.me, show.se = TRUE, show.aic = TRUE, show.dev = TRUE)
```

::: {.callout-tip title="ICC"}
The intraclass correlation coefficient (ICC) "ranges from $0$ if the
grouping conveys no information to $1$ if all members of a group are
identical" [@gelmanDataAnalysisUsing2007: 258]. In other words, it
indicates how much of the variance in the outcome can be explained by
the grouping factor (e.g. school or participant).
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Extract random effects and their standard errors
ranef_obj <- ranef(varmorph.me)  # Extract random effects with conditional variance
se_ranef <- arm::se.ranef(varmorph.me)           # Extract standard errors for random effects

# Prepare a data frame for 'subj_id' random intercepts with confidence intervals
subj_ranef <- ranef_obj$subj_id  # Random effects for subjects
subj_se <- se_ranef$subj_id      # Standard errors for subjects

# Combine random effects and standard errors into a data frame
subj_df <- data.frame(
  subj_id = rownames(subj_ranef),
  intercept = subj_ranef[, "(Intercept)"],
  se = subj_se[, "(Intercept)"],
  conf.low = subj_ranef[, "(Intercept)"] - 1.96 * subj_se[, "(Intercept)"],
  conf.high = subj_ranef[, "(Intercept)"] + 1.96 * subj_se[, "(Intercept)"]
)

# Create the waterfall plot
ggplot(subj_df, aes(x = reorder(subj_id, intercept), y = intercept)) +
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey10") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  coord_flip() +
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey10") +
  labs(title = "Random Intercepts by Subject",
       x = "Subject ID",
       y = "Random Intercept") +
  theme_minimal()

```

#### Varying-slope model

```{r, output = F}
# Varying-slope model; replace 0 with 1 if you want the intercept to vary too
varmorph.me2 <- lmer(rt ~ prime_type +
                      (0 + prime_type | subj_id),
                      data = varmorph2)

summary(varmorph.me2)
```

```{r}
tab_model(varmorph.me2, show.se = TRUE, show.aic = TRUE, show.dev = TRUE)
```

```{r, eval = T, include = T}
#| code-fold: true
#| code-summary: "Show the code"

# Extract random slopes
ranef_data1 <- ranef(varmorph.me2)$subj_id

# Extract standard errors
ranef_data1_se <- arm::se.ranef(varmorph.me2)$subj_id

# Convert data into long format
random_effects_df <- ranef_data1 %>%
  as.data.frame() %>% 
  rownames_to_column(var = "subj_id") %>%
  pivot_longer(cols = -subj_id, 
               names_to = "prime_type", 
               values_to = "random_effect")


# Create a data frame for standard errors
se_df <- as.data.frame(ranef_data1_se) %>%
  rownames_to_column(var = "subj_id") %>%
  pivot_longer(cols = -subj_id, 
               names_to = "prime_type", 
               values_to = "se")


# Combine random effects with standard errors
combined_df <- random_effects_df %>%
  left_join(se_df, by = c("subj_id", "prime_type"))

# Calculate confidence intervals
combined_df <- combined_df %>%
  mutate(lower_ci = random_effect - 1.96 * se,
         upper_ci = random_effect + 1.96 * se)


# Dotplots with confidence intervals
ggplot(combined_df, aes(x = subj_id, y = random_effect, col = prime_type)) +
  coord_flip() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", col = "grey10") +
  facet_wrap(~ prime_type) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Random Effect of Prime Type by Subject ID",
       x = "Random Slope",
       y = "Subject ID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
