---
title: "Ordinal regression"
author:
  name: "Vladimir Buskin" 
  orcid: "0009-0005-5824-1012"
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
format:
  html:
    self-contained: true
    logo: logo.png
    footer: "Regression"
    theme: Reference
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 4
    scrollable: true
editor: visual
bibliography: R.bib
---

## Suggested reading

General:

> @baguleySeriousStatsGuide2012: Chapter 17.4.5
>
> @oconnellLogisticRegressionModels2006
>
> @powersStatisticalMethodsCategorical2008: Chapter 7
>
> [Documentation of Cumulative Link
> Models](https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf)

## Introduction

In her recent contribution, @glassEnglishVerbsCan2021 examines possible
reasons why certain transitive verbs have a stronger affinity towards
object omission compared to others, placing special emphasis on the
routinisation of the actions denoted by the verbs. Specifically, she
assesses **how high/low-routine contexts affect the acceptability of
object omission** for transitive verbs from different frequency bins.

We will replicate her findings using her survey data
`Glass_2021_survey_processed.csv`[^ordinal_regression-1]:

[^ordinal_regression-1]: The original dataset can be retrieved from
    Lelia Glass's OSF repository: <https://osf.io/t6zw5> \[Last
    accessed: 27th September, 2024\].

```{r}
# Load libraries
library(tidyverse)
library(ordinal)
library(MASS)
library(sjPlot)
library(effects)
library(ggeffects)
library(ggpubr)

# For additional tests
library(DescTools)
library(generalhoslem)
library(brant)

# Load data
survey <- read.csv("Glass_2021_survey_processed.csv")

# Inspect dataset
str(survey)
head(survey)
```

::: {.callout-note collapse="true" title="Short breakdown of the variables"}
-   `routine`: In Glass's study, transitive verbs were randomly assigned
    to one of the following conditions:

> -   (High routine condition:) I worked at my poultry farm. Just like I
>     always do, I **butchered** some chickens. Then I gathered some
>     eggs.
>
> -   (Low-routine condition:) I visited a friend's job. Just because
>     people wanted me to try it, I **butchered** some chickens. Then I
>     went for a walk.
>
> Cf. Glass [-@glassEnglishVerbsCan2021: 66]

```{r}
unique(survey$routine)
```

-   `rating` records the responses of participants to a follow-up
    question regarding the acceptability of object omission. The answers
    are recorded on a 1-5 Likert scale.

> *The next time Caroline talks about butchering chickens the day
> before, how likely do you think she is to say the following?*
>
> 'I **butchered** yesterday'
>
> Cf. Glass [-@glassEnglishVerbsCan2021: 66]

```{r}
unique(survey$rating)
```

-   `verb` contains the items to be rated for the conditions in
    `routine`

```{r}
unique(survey$verb)
```

-   `frequency` relates to the frequency bins of the verbs:

```{r}
unique(survey$freq)
```

-   `ParticipantID` identifies each of the 98 subjects who provided
    ratings
:::

## Descriptive overview

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(survey, aes(x = rating, fill = freq)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~freq) +
  theme_minimal() +
  labs(title = "Density of Ratings by Frequency",
       x = "Rating", y = "Density", fill = "Frequency")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(survey, aes(x = rating, fill = routine)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~routine) +
  theme_minimal() +
  labs(title = "Density of Ratings by Routine",
       x = "Rating", y = "Density", fill = "Routine")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

library(ggridges)
ggplot(survey, aes(x = rating, y = verb, fill = verb)) +
  geom_density_ridges(scale = 5, rel_min_height = 0.01, alpha = 0.6) +
  theme_ridges() +
  #theme(legend.position = "none") +
  labs(title = "Distribution of Ratings by Verb",
       x = "Rating", y = "Verb")

```

## Modelling ordinal data

Our task is clear: We need to measure how `routine` and `freq` affect
the variability in the acceptability ratings, while controlling for
repeated measurements for `verb` and `ParticipantID`, which impose a
hierarchical structure on the dataset.

Formally speaking, we have $p$ explanatory variables
$X_1, X_2, ..., X_p$ for $1, ..., p$. The target variable, i.e. our $Y$,
is `rating` with the ordered, discrete outcomes
$y \in \{1, 2, 3, 4, 5\}$.

The goal is to find a model $f$ that describes the relationship between
$Y$ and $X_p$ as accurately as possible and minimises the error term
$\epsilon$:

$$
Y = f(X_1, X_2, ..., X_p) + \epsilon
$$ {#eq-mod}

### Ordered logistic regression

One family of models that respects the ordered, yet categorical nature
of $Y$ is **ordered** (or ordinal) **logistic regression**. Other terms
include **proportional odds models** and **cumulative logit/link
models**.

::: {.callout-note collapse="true" title="Recap: Logistic regression"}
Logistic regression is used to model categorical response variables with
two or more levels. For instance, let's assume our $Y$ is dichotomous
with the following two outcomes:

$$
Y =
\begin{cases}
\text{yes} \\
\text{no}
\end{cases}
$$ {#eq-logreg-outcomes}

Using the logistic function, we can estimate the probability of one
outcome versus the other **given** the predictors $X_p$. Their
log-transformed odds ratio (**log odds**) is equivalent of the
all-too-familiar linear model:

$$
\log\left(\frac{P(Y = yes \mid X_1, X_2, ..., X_p)}{1 - P(Y = yes \mid X_1, X_2, ..., X_p)}\right) = \beta_0 + \sum_{i=1}^p \beta_iX_i
$$ {#eq-logreg}
:::

Core to this approach is the notion of **cumulative probabilities**. Let
$J$ denote the number of ordered categories in $Y$. In Glass's case
study, the estimated cumulative probabilities for each ordered outcome
(= acceptability rating) would have the forms in @eq-cumprobs.

$$
\begin{array}{rcl}
P(Y \leq 1) & = & P(Y = 1) \\
P(Y \leq 2) & = & P(Y = 1) + P(Y = 2) \\
P(Y \leq 3) & = & P(Y = 1) + P(Y = 2) + P(Y = 3) \\
& \vdots & \\
P(Y \leq j) & = & P_1 + ... + P_j
\end{array}
$$ {#eq-cumprobs}

We can now update our logistic regression model to take into account
cumulative probabilities for $j = 1, ..., J-1$.

$$
\log\left(\frac{P(Y \leq j \mid X_1, X_2, ..., X_p)}{1 - P(Y \leq j \mid X_1, X_2, ..., X_p)}\right) = \beta_0 + \sum_{i=1}^p \beta_iX_i
$$ {#eq-cumlog}

The intercepts $\beta_{0_j}$ serve as **cutpoints** between the adjacent
ordinal categories. For $J = 5$ categories, there are $J - 1 = 4$
cutpoints, i.e.,

-   1\|2 for $P(Y \leq 1)$

-   2\|3 for $P(Y \leq 2)$

-   3\|4 for $P(Y \leq 3)$

-   4\|5 for $P(Y \leq 4)$.

Given a change in predictor values, the slope coefficients $\beta_pX_p$
indicate how the probability of being in a higher rating category
changes [@baguleySeriousStatsGuide2012: 691--2].

We can obtain "regular" probabilities from the cumulative ones by
drawing on the equivalence in @eq-cumprobs-transform.

$$
P(Y = j) = P(Y \leq j) - P(Y \leq j - 1)
$$ {#eq-cumprobs-transform}

For instance, the probability $P(Y = 3)$ is equivalent to

$$
P(Y = 3) = P(Y \leq 3) - P(Y \leq 2).
$$ {#eq-example}

::: {.callout-important title="Assumptions of proportional odds models" collapse="false"}
**The proportional odds assumption** stipulates a stable effect of the
predictors on the (log) odds of the ordinal outcomes across all possible
cutpoints [@oconnellLogisticRegressionModels2006: 29]. In case of
violation, it is better to rely on partial proportional odds models or
multinomial logistic regression instead.
:::

## Application in R

There are several R packages that support ordinal logistic regression
models. This section provides an overview of some of the more common (as
well as well-documented) implementations.

### Using `polr()` from the `MASS` library

```{r, echo = TRUE, output = TRUE}
# Convert to factor and determine ordering
survey$rating <- factor(survey$rating, ordered = TRUE, levels = c("1", "2", "3", "4", "5"))

# Fit polr model
survey.polr <- polr(rating ~ 
                      freq +
                      routine,
                      data = survey)

# Model summary
summary(survey.polr)

# R-squared and AIC
PseudoR2(survey.polr, c("Nagelkerke", "AIC"))

```

```{r}
tab_model(survey.polr, show.se = TRUE, show.aic = TRUE, show.dev = TRUE, transform = NULL)
```

::: {.callout-tip title="Interpreting the model parameters" collapse="true"}
-   **Coefficients**: The conditions `freqlo` (low frequency) and
    `routinelo` (low-routine context) both have negative values, which
    means that both of them decrease the probability of obtaining a
    higher acceptability rating (compared to `freqhi` and `routinehi`).

-   **Intercepts**: These represent the cutpoints between the ordinal
    categories, which are necessary for calculating the probabilities of
    each ordinal category.
:::

### Testing assumptions and goodness of fit

-   Test proportional odds assumption:

```{r}
brant(survey.polr) # p < 0.05 is a violation of the assumption
```

-   Hosmer-Lemeshow test, which is essentially a $\chi^2$-test:

```{r}
logitgof(survey$rating, # observed
         fitted(survey.polr), # expected
         ord = TRUE) # respect ordering
```

-   The Lipsitz test is an extension of the Hosmer-Lemeshow test. Note
    that it requires the response to be a factor.

```{r}
lipsitz.test(survey.polr)
```

-   Part of the same family of tests is the Pulkstenis-Robinson test,
    which also relies on the $\chi^2$-distribution:

```{r}
pulkrob.chisq(survey.polr, catvars = c("freq", "routine"))
```

### Visualisation

#### With `effects`

```{r, output = F}
# Routine effect plot
plot(Effect(focal.predictors = c("routine"), mod = survey.polr), rug = FALSE, style="stacked")

```

#### With `ggeffects` and `ggplot2`

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Show the code"

# Get the ggeffects data
eff <- ggeffects::ggeffect(survey.polr, "freq")

# Convert to a data frame
plot_data <- as.data.frame(eff)

# Ensure the response.level has the desired levels
plot_data$response.level <- factor(plot_data$response.level, 
                                   levels = c("X1", "X2", "X3", "X4", "X5"),
                                   labels = c("1", "2", "3", "4", "5"))

# Create the plot with confidence intervals
p1 <- ggplot(plot_data, aes(x = x, y = predicted, color = response.level, group = response.level)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.5) +
  scale_color_viridis_d() +
  labs(
    x = "Frequency",
    y = "Predicted Probability",
    color = "Rating"
  ) +
  facet_grid(~ response.level) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(margin = margin(b = 10))
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

p1


```

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Show the code"

# Get the ggeffects data for "routine"
eff_routine <- ggeffects::ggeffect(survey.polr, "routine")

# Convert to a data frame
plot_data_routine <- as.data.frame(eff_routine)

# Ensure the response.level has the desired levels for "routine"
plot_data_routine$response.level <- factor(plot_data_routine$response.level, 
                                           levels = c("X1", "X2", "X3", "X4", "X5"),
                                           labels = c("1", "2", "3", "4", "5"))

# Create the second plot for "routine"
p2 <- ggplot(plot_data_routine, aes(x = x, y = predicted, color = response.level, group = response.level)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.5) +
  scale_color_viridis_d() +
  labs(
    x = "Routine",
    y = "Predicted Probability",
    color = "Rating"
  ) +
  facet_grid(~ response.level) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(margin = margin(b = 10))
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

p2
```

```{r, echo = TRUE, output = TRUE}
#| code-fold: true
#| code-summary: "Show the code"

# Generate the interaction effects for "freq" and "routine"
eff_interaction <- ggeffect(survey.polr, terms = c("freq", "routine"))

# Convert to a data frame
plot_data_interaction <- as.data.frame(eff_interaction)

# Ensure the response.level has the desired levels
plot_data_interaction$response.level <- factor(plot_data_interaction$response.level, 
                                               levels = c("X1", "X2", "X3", "X4", "X5"),
                                               labels = c("1", "2", "3", "4", "5"))


# Create the interaction plot with facet by 'x' and color by 'response.level'
p_interaction <- ggplot(plot_data_interaction, aes(x = group, y = predicted, color = response.level, group = response.level)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.5) +
  scale_color_viridis_d() +
  labs(
    title = "Predicted Probabilities for Interaction of Frequency and Routine",
    x = "Routine",
    y = "Predicted Probability",
    color = "Rating"
  ) +
  facet_wrap(~ x, labeller = labeller(x = c("hi" = "High Frequency", "lo" = "Low Frequency"))) +  # Facet by "freq"
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(margin = margin(b = 10))
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Display the interaction plot
p_interaction


```

### Using `clm()` from the `ordinal` library

```{r, echo = T, output = F}
# Convert to factor and determine ordering
survey$rating <- factor(survey$rating, ordered = TRUE, levels = c("1", "2", "3", "4", "5"))

# Fit cumulative link model
clm.1 <- ordinal::clm(rating ~ 
                    freq +
                    routine,
                    data = survey, Hess=TRUE)

# Model summary
summary(clm.1)
```

```{r}
tab_model(clm.1, show.se = TRUE, show.aic = TRUE, show.dev = TRUE, transform = NULL)
```

### Mixed-effects ordinal regression

::: {.callout-note collapse="true" title="Recap: Mixed-effects models"}
If the data is nested according to some grouping factor with $1, ..., k$
groups, we can let the intercept and/or slopes vary by group. For
instance, recall the varying-intercept model:

$$
Y = \alpha_{k} + \beta_1X_{1} + \beta_2X_{2} + ... + \epsilon \qquad \alpha_{k} \sim N(\mu_{\alpha}, \sigma_{\alpha}^2).
$$ In this case we also speak of **random effects**.
:::

```{r, echo = T, output = T}
# Convert to factor and determine ordering
survey$rating <- factor(survey$rating, ordered = TRUE, levels = c("1", "2", "3", "4", "5"))

# Fit mixed model with random intercepts for "verb" and "ParticipantID"
clm.2 <- ordinal::clmm(rating ~ 
                    freq * routine +
                    (1 | verb) +
                    (1 | ParticipantID),
                    data = survey, Hess=TRUE)

# Model summary
summary(clm.2)
```

```{r}
tab_model(clm.2, show.se = TRUE, show.aic = TRUE, show.dev = TRUE, transform = NULL)
```

```{r, echo = T, output = T}
#| code-fold: true
#| code-summary: "Show the code"

# Extract random effects
re_verb <- ranef(clm.2)$verb
re_participant <- ranef(clm.2)$ParticipantID

# Create dataframes for random effects
df_verb <- data.frame(verb = rownames(re_verb), re = re_verb[,1])
df_participant <- data.frame(ParticipantID = rownames(re_participant), re = re_participant[,1])

# Get predictions for an average case
pred_avg <- ggpredict(clm.2, terms = c("freq", "routine"))

# Add random effects to predictions
pred_verb <- crossing(pred_avg, df_verb) %>%
  mutate(predicted = predicted + re)

pred_participant <- crossing(pred_avg, df_participant) %>%
  mutate(predicted = predicted + re)

# Create a horizontal dot plot for random effects of participants
p_caterpillar <- ggplot(df_participant, aes(x = re, y = reorder(ParticipantID, re))) +
  geom_point(size = 3, color = "steelblue3") +  # Dots representing the random effects
  labs(title = "Random Effects for Participants", 
       x = "Random Effect Estimate (log odds)", 
       y = "Participant ID") +
   geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())  # Removes the gridlines for y-axis

p_caterpillar2 <- ggplot(df_verb, aes(x = re, y = reorder(verb, re))) +
  geom_point(size = 3, color = "steelblue3") +  # Dots representing the random effects
  labs(title = "Random Effects for Verbs", 
       x = "Random Effect Estimate (log odds)", 
       y = "Verb") +
   geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())  # Removes the gridlines for y-axis

ggarrange(p_caterpillar, p_caterpillar2, ncol = 2, common.legend = TRUE, legend = "right")

```

## Generalised Additive Mixed-effects Models (GAMMs)

### Suggested reading

For linguists:

> Baayen & Linke [-@baayenGeneralizedAdditiveMixed2020]

General:

> Hastie & Tibshirani [-@hastieGeneralizedAdditiveModels1991]
>
> Wood [-@woodGeneralizedAdditiveModels2006]

### Rationale

A core assumption of Generalised Linear Models (GLMs) is a linear
relationship between predictor(s) and response. If, however, one is
interested in exploring potential non-linear trends without the risk of
extreme overfitting, GAMs offer an elegant solution: Instead of relying
on the linear sum of model coefficients, GAMs estimate more flexible
**smooth terms** $f_k$ for $k = 1, ..., p$. For illustration, @eq-gam
shows a linear additive model for a continuous target variable with $p$
predictors.

$$
Y = \beta_0 + \sum\limits_{k = 1}^p f_k(X_k)
$$ {#eq-gam}

### Application in R

```{r}
# Load libraries
library(mgcv)
library(itsadug)
library(gratia)

# Convert predictors to factors
survey$ParticipantID <- as.factor(survey$ParticipantID)
survey$verb <- as.factor(survey$verb)

# Fit GAMM
gam1 <- bam(as.numeric(rating) ~ # treated as numeric term
              freq + # linear term
              routine + # linear term
              s(ParticipantID, bs = "re") + # smooth term
              s(verb, bs = "re"), # smooth term
              data = survey, 
              family = ocat(R = 5) # number of ordinal categories
            )

# Model summary
summary(gam1)

# Extract the intercepts for plotting
thresh <- gratia::theta(gam1) %>% 
  tibble::as_tibble() %>% 
  setNames(c("threshold"))

# Extract predictions for "routine"
routine_pred <- ggpredict(gam1, terms = "routine")

# Plot predictions
routine_pred %>% 
  ggplot(aes(x = x, y = predicted)) +
  geom_point(col = "steelblue") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, width = 0), col = "steelblue") +
  geom_hline(data = thresh, aes(yintercept = threshold), linetype = "dashed") +
  theme_minimal()


# Extract random effects for "verb"
verb_pred <- ggpredict(gam1, terms = "verb")

# Plot random effect
verb_pred %>% 
  ggplot(aes(x = x, y = predicted)) +
  geom_point(col = "steelblue") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, width = 0), col = "steelblue") +
  geom_hline(data = thresh, aes(yintercept = threshold), linetype = "dashed") +
  theme_minimal()


# Extract random effects for "ParticipantID"
subj_pred <- ggpredict(gam1, terms = "ParticipantID")

# Plot random effect
subj_pred |>
  ggplot(aes(x = x, y = predicted)) +
  geom_point(col = "steelblue") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, width = 0), col = "steelblue") +
  #geom_line() +
  geom_hline(data = thresh, aes(yintercept = threshold), linetype = "dashed") +
  theme_minimal()

```
