---
title: "Principal Components Analysis"
author: Vladimir Buskin
format:
  html:
    self-contained: true
    code-fold: false
    theme: default
    toc: true
    toc-depth: 4
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
bibliography: R.bib
metadata-files:
  - _quarto.yml
google-scholar: false
---

## Recommended reading

> Mair [-@mairModernPsychometrics2018: Chapter 6]

## Preparation

This unit relies on psycholinguistic data from the [South Carolina
Psycholinguistic
Metabase](https://sc.edu/study/colleges_schools/artsandsciences/psychology/research_clinical_facilities/scope/)
[@gaoSCOPESouthCarolina2022].[^pca-1] Detailed descriptions of the
variables can be found
[here](https://sc.edu/study/colleges_schools/artsandsciences/psychology/research_clinical_facilities/scope/search.php).

[^pca-1]: One exception is the variable
    `Resnik_strength [`@resnikSelectionalConstraintsInformationtheoretic1996\],
    which was computed manually and appended to the data frame.

The data frame `scope_sem_df` contains semantic ratings for a sample of
1,702 transitive verbs. Note that all columns have been standardised
(cf. `?scale()` for details).

```{r}
# Load libraries
library(tidyverse)
library(purrr)
library(lattice)
library(corrplot)
library(psych)
library(GPArotation)
library(Gifi)


# Load data
scope_sem_df <- readRDS("scope_sem_df.RDS")

# Overview
glimpse(scope_sem_df)
```

## Examine correlations

```{r}
# Remove the verb labels from the first column
scope_sem_df_cleaned <- scope_sem_df[,-1]

# Generate correlation matrix
cor_mat1 <- cor(scope_sem_df_cleaned)

# Plot correlation matrix
corrplot(cor_mat1, col = topo.colors(200), tl.col = "darkgrey", number.cex = 0.5, tl.cex = 0.5)
```

```{r}
# Levelplot
seq1 <- seq(-1, 1, by = 0.01)

levelplot(cor_mat1, aspect = "fill", col.regions = topo.colors(length(seq1)),
          at = seq1, scales = list(x = list(rot = 45)),
          xlab = "", ylab = "")

```

Needless to say, the above correlation matrices are hard to interpret –
even more so if the number of variables were to increase further.

**Principal Components Analysis** offers a technique to break down a
high-dimensional dataset, into a much smaller set of "meta-variables", i.e., **principle components** (PCs) which capture the bulk of the variance in the data. This allows researchers to see overarching patterns in the data and re-use the output for further analysis
(e.g., clustering or predictive modelling).

## How does PCA work?

PCA "repackages" large sets of variables by forming uncorrelated linear combinations of them, yielding $k$ principal components $Z_k$ of the dataset (for $k \in \{1, ..., K\}$).

Each PC comprises a set of **loadings** (or **weights**) $w_{nm}$, which are comparable to the coefficients of regression equations. For instance, the first PC has the general form shown in the equation below, where $x_m$ stand for continuous input variables in the $n \times m$ data matrix.

\begin{equation} \label{pca-1}
Z_1 = w_{11}\mathbf{x}_{1} + w_{21}\mathbf{x}_{2} + \dots + w_{m1}\mathbf{x}_{m}
\end{equation}

If a feature positively loads on a principal component (i.e., $w > 0$), it means that as the value of this feature increases, the score for this principal component also increases. The magnitude of $w$ indicates the strength of this relationship. Conversely, negative loadings ($w < 0$) indicate that as the feature value increases, the PC score decreases as well.

The figure below offers a visual summary of PCA:


```{r, echo = FALSE, output = FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

# Generate sample data
set.seed(42)
n <- 100
x <- rnorm(n)
y <- 0.5 * x + rnorm(n, sd = 0.5)
df <- data.frame(x = x, y = y)

# Function to calculate principal components
calculate_pca <- function(df) {
  pca <- prcomp(df, center = TRUE, scale. = TRUE)
  pc1 <- pca$rotation[, 1]
  pc2 <- pca$rotation[, 2]
  list(pc1 = pc1, pc2 = pc2, center = pca$center)
}

# Calculate PCA
pca_result <- calculate_pca(df)

# Function to create plot
create_pca_plot <- function(df, pc1, pc2, center, title, show_pc1 = FALSE, show_pc2 = FALSE) {
  p <- ggplot(df, aes(x, y)) +
    geom_point(alpha = 0.6) +
    labs(title = title) +
    theme_minimal() +
    coord_fixed(ratio = 1)
  
  if (show_pc1) {
    p <- p + geom_segment(aes(x = center[1], y = center[2], 
                     xend = center[1] + pc1[1], 
                     yend = center[2] + pc1[2]),
                 arrow = arrow(length = unit(0.2, "cm")), 
                 color = "red", size = 1)
  }
  
  if (show_pc2) {
    p <- p + geom_segment(aes(x = center[1], y = center[2], 
                              xend = center[1] + pc2[1], 
                              yend = center[2] + pc2[2]),
                          arrow = arrow(length = unit(0.2, "cm")), 
                          color = "blue", size = 1)
  }
  
  p
}

# Create plots
p1 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "Original Data")

p2 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "First Principal Component (PC1)", show_pc1 = TRUE)

p3 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "PC1 and PC2", show_pc1 = TRUE, show_pc2 = TRUE)

# Function to project data onto PC1
project_data <- function(df, pc, center) {
  df_centered <- df - center
  projected <- as.matrix(df_centered) %*% pc
  data.frame(PC = projected, original_x = df$x, original_y = df$y)
}

# Project data onto PC1
df_projected <- project_data(df, pca_result$pc1, pca_result$center)

# Create plot with data colored by PC1 score
p4 <- ggplot(df_projected, aes(x = original_x, y = original_y, color = PC)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Direction of variance captured by PC1",
       x = "x", y = "y", color = "PC1 Score") +
  theme_minimal()

# Display plots
grid.arrange(p1, p2, p3, p4, ncol = 2)
```


```{r, echo = FALSE}
# Generate sample data
set.seed(42)
n <- 100
x <- rnorm(n)
y <- 0.5 * x + rnorm(n, sd = 0.5)
df <- data.frame(x = x, y = y)

# Function to calculate principal components
calculate_pca <- function(df) {
  pca <- prcomp(df, center = TRUE, scale. = TRUE)
  pc1 <- pca$rotation[, 1]
  pc2 <- pca$rotation[, 2]
  list(pc1 = pc1, pc2 = pc2, center = pca$center, pca = pca)
}

# Calculate PCA
pca_result <- calculate_pca(df)

# Function to create plot
create_pca_plot <- function(df, pc1, pc2, center, title, show_pc1 = FALSE, show_pc2 = FALSE) {
  p <- ggplot(df, aes(x, y)) +
    geom_point(alpha = 0.6) +
    labs(title = title) +
    theme_minimal() +
    coord_fixed(ratio = 1)
  
  if (show_pc1) {
    p <- p + geom_segment(aes(x = center[1], y = center[2], 
                     xend = center[1] + pc1[1], 
                     yend = center[2] + pc1[2]),
                 arrow = arrow(length = unit(0.2, "cm")), 
                 color = "red", size = 1)
  }
  
  if (show_pc2) {
    p <- p + geom_segment(aes(x = center[1], y = center[2], 
                              xend = center[1] + pc2[1], 
                              yend = center[2] + pc2[2]),
                          arrow = arrow(length = unit(0.2, "cm")), 
                          color = "blue", size = 1)
  }
  
  p
}

# Create plots
p1 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "Original Data")

p2 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "PC1", show_pc1 = TRUE)

p3 <- create_pca_plot(df, pca_result$pc1, pca_result$pc2, pca_result$center, 
                      "PC2", show_pc1 = TRUE, show_pc2 = TRUE)

# Function to project data onto PCs
project_data <- function(df, pca) {
  as.data.frame(predict(pca, df))
}

# Project data onto PCs
df_projected <- project_data(df, pca_result$pca)
df_projected$original_x <- df$x
df_projected$original_y <- df$y

# Create plot with data colored by PC1 score
p4 <- ggplot(df_projected, aes(x = original_x, y = original_y, color = PC1)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Direction of variance captured by PC1",
       x = "x", y = "y", color = "PC1 Score") +
  theme_minimal()

# Create plot with data colored by PC2 score
p5 <- ggplot(df_projected, aes(x = original_x, y = original_y, color = PC2)) +
  geom_point() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Direction of variance captured by PC2",
       x = "x", y = "y", color = "PC2 Score") +
  theme_minimal()

# Display plots
grid.arrange(p2, p3, p4, p5, ncol = 2)
```



::: {.callout-tip title="How do you identify the principle components of a dataset?" collapse="true"}

The computation of PC loadings/weights relies on several techniques from matrix algebra, namely **singular value decomposition** (SVD) and **eigenvalue decomposition**, depending on the format of the input matrix.

Let $\mathbf{X}$ denote the $n \times m$ matrix of input variables. Using SVD, it can be analysed into three elements:

$$
\mathbf{X} = \mathbf{UDV}^T
$$


```{r, eval = F, include = F}
library("mvtnorm")
sigma <- matrix(c(2, 0.8, 0, 0.8, 0.5, 0, 0, 0, 0.1), ncol = 3)
set.seed(123)
baguette <- rmvnorm(1000, mean = c(0,0,0), sigma = sigma)

svdb <- svd(baguette)

svdb$u
D <- diag(svdb$d)   ## singular values
round(D, 3)
V <- svdb$v   


v1 <- V[,1]
v2 <- V[,2]
round(v1 %*% v2, 7)  


library("rgl")
plot3d(baguette, col = "gray", xlim = c(-4, 4), ylim = c(-4,4),
         zlim = c(-4,4), aspect = 1, size = 2)
  arrow3d(c(0,0,0), V[,1], col = "red")
  arrow3d(c(0,0,0), V[,2], col = "red")
  arrow3d(c(0,0,0), V[,3], col = "red")
```


:::


