---
title: "Poisson regression"
author:
  name: "Vladimir Buskin" 
  orcid: "0009-0005-5824-1012"
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
format:
  html:
    self-contained: true
    logo: logo.png
    footer: "Regression"
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 4
    scrollable: true
editor: visual
bibliography: R.bib
---

## Recommended reading

> @winter_statistics_2020: Chapter 13
>
> @baguleySeriousStatsGuide2012: Chapter 17.5

## Preparation

```{r}
# Load libraries
library(tidyverse)
library(ggeffects)
library(readxl)
library(car)

# Load datasets
verbs <- read_xlsx("winter_2020_visual.xlsx")
```

## The Poisson family

Frequency data is ubiquitous in corpus linguistics. Given its numeric
nature, it seems tempting to model such data using [linear
regression](Linear_regression.qmd), but doing so is bound to cause
problems. Partially owing to the fact that count data is always
positive, the residuals more often than not deviate from normality and
additionally display non-constant variance. The figures below illustrate
these issues for a linear model fitted to simulated count data.

```{r, echo = F}
set.seed(123)

# Simulate predictor variable (continuous)
n <- 500
x <- runif(n, 0, 10)  # Random values between 0 and 10

# Generate Poisson-distributed count data
lambda <- exp(0.5 + 0.3 * x)  # Log-link relationship
y <- rpois(n, lambda)  # Poisson-distributed response variable

# Fit a linear regression model
lm_fit <- lm(y ~ x)

# Plot residuals to examine non-normality and heteroscedasticity
par(mfrow = c(1, 2))  # Arrange plots side by side


crPlot(lm_fit, var = "x") 

plot(lm_fit, which = 1)

```

A probability distribution that is much better equipped for this special
case of discrete, yet numeric data is the **Poisson distribution**.
Assuming a Poisson-distributed variable $X$, its exact shape is
determined by a single parameter, $\lambda$ (lambda), which is both its
mean **and** and variance. In other words,

$$
X \sim Pois(\lambda).
$$ {#eq-pois}

Its probability mass function has a notable negative skew, which becomes
less conspicuous for increasingly higher parameter values.

```{r, echo = F}

x1 <- dpois(0:20, 3.5)
x2 <- dpois(0:20, 10)

par(mfrow = c(1, 2))

barplot(x1, names.arg = 0:20, xlab = "Count", ylab = "Probability", main = "Poisson pmf with lambda = 3.5")

barplot(x2, names.arg = 0:20, xlab = "Count", ylab = "Probability", main = "Poisson pmf with lambda = 10")

```

## Poisson regression

Recall the linear regression @eq-multreg, where the dependent variable
$Y$ was modelled as a function of the linear sum of predictor terms
$\beta_pX_p$ for $p$ independent variable.

The **Poisson model** is very similar to the linear regression model,
with the main differences being the logarithmic transformation of the
response and the exclusion of the error term:

$$ \ln(Y) = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_pX_p.
$$ {#eq-pois-full}

To remove the logarithm and obtain the model output on a more intuitive
scale, we simply exponentiate both sides:

$$ Y = e^{\beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_pX_p}.
$$ {#eq-pois-exp}

### Application in R

The data provided by Winter [-@winter_statistics_2020] contains
frequency data for hundreds of verbs (`Freq` column) as well as a
variety of psycholinguistic ratings (`Sight`, `Touch`, `Sound` etc.),
which were originally compiled by Lynott & Connell
[-@LynottConnell2009].

```{r}
head(verbs)
```

We will examine of the effects of different sensory ratings on the
frequency of a verb. When fitting the generalized linear model, it is
important to indicate the argument `family = "poisson"` to apply the
correct (logarithmic) link function.

```{r}
# Fit Poisson regression model
freq.m1 <- glm(Freq ~ Sight + Touch + Sound + Taste + Smell, data = verbs, family = "poisson")

# Summarise model statistics
summary(freq.m1)
```

The model has identified numerous significant effects, i.e.,
$\beta$-values that are significantly different from 0. The low standard
errors hint at very robust estimates, resulting in 95% confidence
intervals that are barely visible in the effect plots.

```{r}
# Get predicted values for each predictor
pred_sight <- ggpredict(freq.m1, terms = "Sight")
pred_touch <- ggpredict(freq.m1, terms = "Touch")
pred_sound <- ggpredict(freq.m1, terms = "Sound")
pred_taste <- ggpredict(freq.m1, terms = "Taste")
pred_smell <- ggpredict(freq.m1, terms = "Smell")

# For plotting individual predictions, simply use:
# plot(pred_sight) 

# Combine all predictions into one data frame
pred_all <- rbind(
  data.frame(pred_sight, predictor = "Sight"),
  data.frame(pred_touch, predictor = "Touch"),
  data.frame(pred_sound, predictor = "Sound"),
  data.frame(pred_taste, predictor = "Taste"),
  data.frame(pred_smell, predictor = "Smell")
)

# Create faceted plot
ggplot(pred_all, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.5) +
  facet_wrap(~predictor, scales = "free_x") +
  labs(x = "Rating", y = "Predicted verb frequency") +
  theme_bw()
```

### Interpretation

As we can see, verbs with higher `Sight` ratings (i.e., highly visual
verbs) are associated with the greatest increase in frequency of
occurrence. Given a one-unit increase in $X$, we can obtain the
percentage increase (or decrease, respectively) using the formula
$100(e^b - 1)$. The predictor `Sight` has an estimate of approximately
$0.87$ ($p< 0.001$, 95% CI \[0.87, 0.88\]), so the proportional increase
is $100(e^{0.87} - 1) = 138.69\%$.

By contrast, `Smell` is a associated with a lower frequency:
$100(e^{0.10} -1) = -9.51\%$ is the drop in frequency for each one-unit
increase in smell ratings. In essence: Smelly verbs are unpopular!
