---
title: "Regular expressions"
author: Vladimir Buskin, Thomas Brunner
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
    
editor: visual
---

## Regular expressions

**Regular expressions** (or 'regex') help us find more complex patterns
in strings of text. Suppose we are interested in finding all
inflectional forms of the lemma PROVIDE in a corpus, i.e., *provide,
provides, providing* and *provided*. Insteading of searching for all
forms individually, we can construct a regular expression of the form

$$
\text{provid(es | ing | ed)?}
$$which can be read as 'Match the sequence of letters \<provide\> as
well as when it is optionally followed by the letters \<s\> or \<ing\>
or \<ed\>'. Notice how optionality is signified by the '?' operator and
alternatives by '\|'.

To activate regular expression in a KWIC query, simply set the
`valuetype` argument to `"regex"`:

```{r, eval = TRUE}

# Load library and corpus
library(quanteda)
ICE_GB <- readRDS("ICE_GB.RDS")

# Perform query
kwic_provide <- kwic(ICE_GB,
                     phrase("provid(es|ing|ed)?"),
                     valuetype = "regex",
                     window = 20)
```

The number of hits has more than doubled. However, upon closer
inspection, we'll notice a few false positives, namely *providential,
provider* and *providers*:

```{r}
table(kwic_provide$keyword)
```

There are two ways to handle this:

1.  Refine the search expression further to only match those cases of
    interest.
2.  Manually sort out irrelevant cases during annotation in your
    spreadsheet software.

As a rule of thumb, you should consider improving your search expression
if you receive hundreds or even thousands of false hits. If there are
only a couple of false positives, it's usually easier to simply mark
them as "irrelevant" in your spreadsheet.

::: callout-caution
## Task

How could you refine the search expression for PROVIDE to get rid of the
irrelevant cases? Consult the RegEX Cheatsheet below!

```{r, output = FALSE}
#| code-fold: true
#| code-summary: "Solution:"

# Add word boundary with \\b
kwic_provide2 <- kwic(ICE_GB,
                     phrase("\\bprovid(e|es|ing|ed)\\b"),
                     valuetype = "regex",
                     window = 20)

table(kwic_provide2$keyword)

```
:::

## A RegEx Cheatsheet

### Basic functions

| **Command** | **Definition** | **Example** | **Finds**           |
|-------------|----------------|-------------|---------------------|
|             |                | `python`    | *python*            |
| `.`         | Any character  | `.ython`    | *aython, bython...* |

### Character classes and alternatives

| **Command** | **Definition**                               | **Example**     | **Finds**                       |
|-----------------|---------------------|-----------------|-----------------|
| `[abc]`     | Class of characters                          | `[jp]ython`     | *jython, python*                |
| `[ ^pP]`    | Excluded class of characters                 | `[^pP]ython`    | everything but *python, Python* |
| `(...|...)` | Alternatives linked by logical operator `or` | `P(ython|eter)` | *Python, Peter*                 |

### Pre-defined character classes

| **Command** | **Definition**                           | **Example**    | **Finds**                        |
|-----------------|--------------------|-----------------|-----------------|
| `\\w`       | All alphanumeric characters              |                | A-Z, a-z, 0-9                    |
| `\\W`       | All non-alphanumeric characters          |                | everything but A-Z, a-z, 0-9     |
| `\\d`       | All decimal numbers                      |                | 0-9                              |
| `\\D`       | Everything which is not a decimal number |                | everything but 0-9               |
| `\\s`       | Empty space                              |                |                                  |
| `\\b`       | Word boundary                            | `\\bpython\\b` | Matches *python* as a whole word |

### Quantifiers

| **Command** | **Definition**                                | **Example**   | **Finds**                     |
|-----------------|---------------------|-----------------|-----------------|
| `?`         | One or zero instances of the preceding symbol | `Py?thon`     | *Python, Pthon*               |
| `*`         | No matter how many times — also zero          | `Py*thon`     | *Python, Pthon, Pyyyython...* |
|             |                                               | `P[Yy]*thon`  | *Python, Pthon, PyYYython...* |
| `+`         | No matter how many times but at least once    | `Py+thon`     | *Python, Pyyython, Pyyyython* |
| `{1,3}`     | `{min, max}`                                  | `Py{1,3}thon` | *Python, Pyython, Pyyython*   |

## Exercises

1.  Find all labels of months!
2.  Write an elegant regular expression which finds *sing*, *sang* and
    *sung*.
3.  Find all four-digit numbers in the corpus!
4.  Write an elegant regular expression which finds all inflectional
    forms of *swim*!
