---
title: "Regular expressions"
author: Vladimir Buskin, Thomas Brunner
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
    
editor: visual
---

## Regular expressions

**Regular expressions** (or 'regex') help us find more complex patterns
in strings of text. Suppose we are interested in finding all
inflectional forms of the lemma PROVIDE in a corpus, i.e., *provide,
provides, providing* and *provided*. Insteading of searching for all
forms individually, we can construct a regular expression of the form

$$
\text{provide(s | ing | ed)?}
$$ which can be read as 'Match the sequence of letters \<provide\> as
well as when it is optionally followed by the letters \<s\> or \<ing\>
or \<ed\>'. Notice how optionality is signified by the '?' operator and
alternatives by '\|'.

To activate regular expression in a KWIC query, simply set the
`valuetype` argument to `"regex"`:

```{r, eval = FALSE}

# Query the corpus
kwic_provide3 <- kwic(ICE_GB,
                     pattern = "provide(s|ing|ed)?",
                     valuetype = "regex", # query format
                     window = 20)

                     

# View output in separate window
#View(kwic_provide)
```

The number of hits has more than doubled. However, upon closer
inspection, we'll notice a number of false positives (e.g.,
*providential*).

```{r, echo = F, eval = FALSE}

kwic_provide3 %>% 
  head() %>% 
  kbl() %>% 
  kable_classic(full_width = T, html_font = "Cambria")

```

## A RegEx Cheatsheet

### Basic functions

| **Command** | **Definition** | **Example** | **Finds** |
|-------------|----------------|-------------|-----------|
|             |                | `python`    | *python*  |
| `.`         | Any character  | `.ython`    | *aython, bython...* |

### Character classes and alternatives

| **Command**          | **Definition**                                     | **Example**   | **Finds**                 |
|----------------------|---------------------------------------------------|---------------|---------------------------|
| `[abc]`              | Class of characters                               | `[jp]ython`   | *jython, python*          |
| `[ ^pP]`             | Excluded class of characters                      | `[^pP]ython`  | everything but *python, Python* |
| `(...|...)`          | Alternatives linked by logical operator `or`      | `P(ython|eter)` | *Python, Peter*           |

### Pre-defined character classes

| **Command**  | **Definition**                   | **Example** | **Finds**               |
|--------------|----------------------------------|-------------|-------------------------|
| `\w`         | All alphanumeric characters      |             | A-Z, a-z, 0-9           |
| `\W`         | All non-alphanumeric characters  |             | everything but A-Z, a-z, 0-9 |
| `\d`         | All decimal numbers              |             | 0-9                     |
| `\D`         | Everything which is not a decimal number |       | everything but 0-9      |
| `\s`         | Empty space                      |             |                         |

### Quantifiers

| **Command**  | **Definition**                                             | **Example**    | **Finds**                              |
|--------------|------------------------------------------------------------|----------------|----------------------------------------|
| `?`          | One or zero instances of the preceding symbol              | `Py?thon`      | *Python, Pthon*                        |
| `*`          | No matter how many times — also zero                       | `Py*thon`      | *Python, Pthon, Pyyyython...*          |
|              |                                                            | `P[Yy]*thon`   | *Python, Pthon, PyYYython...*          |
| `+`          | No matter how many times but at least once                 | `Py+thon`      | *Python, Pyyython, Pyyyython*          |
| `{1,3}`      | `{min, max}`                                               | `Py{1,3}thon`  | *Python, Pyython, Pyyython*            |

## Tasks

1. Find all labels of months!
2. Write an elegant regular expression which finds *sing*, *sang* and *sung*.
3. Find all four-digit numbers in the corpus!
4. Write an elegant regular expression which finds all inflectional forms of *swim*!