---
title: "Continuous data"
author: Vladimir Buskin
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
bibliography: R.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

## Preparation

We will use the dataset from the previous unit:

```{r, echo = TRUE, output = FALSE}

# Libraries
library("readxl")
library("tidyverse")

# Load data from working directory
cl.order <- read_xlsx("Paquot_Larsson_2020_data.xlsx")

```

## Measures of central tendency {.scrollable}

### The mean {.smaller .scrollable}

A useful summary statistic is the arithmetic mean $\bar{x}$ [cf.
@heumann_introduction_2022: 38]. Consider a continuous variable $X$ with
observations $x_1, x_2, ..., x_n$ from a sample of size $n$. The sample
mean $\bar{x}$ then corresponds to

$$
\bar{x} = \frac{x_1 + x_2 + ... + x_n}{n} \\ = \frac{1}{n}\sum_{i=1}^n{x_i}.
$$

::: callout-note
## Hold on, what does this large Σ symbol mean?

The summation symbol $\sum$ is a concise way to express the sum of a
sequence of numbers. It works like this:

-   The expression below the $\sum$ (e.g., $i = 1$) indicates the
    starting value of the index $i$.

-   The number above the $\sum$ (e.g., $n$) is the ending value of the
    index $i$.

-   The expression to the right of the $\sum$ (e.g., $x_i$) is the term
    to be summed.

For instance, the expression $\sum_{i=1}^n{x_i}$ means "sum the values
of $x_i$ starting from $i = 1$ up to $i = n$." In other words, it adds
up the values $x_1 + x_2 + \dots + x_n$.
:::

In R, we can obtain the average value of a numeric vector with the
`mean()` function. Let's do that for the length of main clauses found in
the `cl.order` data:

```{r, echo = TRUE, output = TRUE}
mean(cl.order$LEN_MC)
```

The output returned by this function provides a one-value summary of all
observations contained in `LEN_MC`. How could we visualise it?

First, let's create a histogram of `LEN_MC`. Histograms summarise how
often certain values occur in a continuous variable. The blue line
indicates the sample mean. Alternatively, you can use a density plot to
represent the probability of encountering a specific value in the data.

::: panel-tabset
### Histogram (ggplot2)

```{r, echo = TRUE, warnings = FALSE, message = FALSE}

# Plot distribution of LEN_MC
cl.length.hist <- ggplot(cl.order, aes(x = LEN_MC)) +
                  geom_histogram(binwidth = 2)

cl.length.hist +
  # Add mean
  geom_vline(aes(xintercept = mean(LEN_MC)),
             color = "steelblue",
             linewidth = 1) +
  theme_classic()
  
```

### Density plot (ggplot2)

```{r, echo = TRUE, warnings = FALSE, message = FALSE}

# Plot distribution of LEN_MC
cl.length.dens <- ggplot(cl.order, aes(x = LEN_MC)) +
                  geom_density()

cl.length.dens +
  # Add mean
  geom_vline(aes(xintercept = mean(LEN_MC)),
             color = "steelblue",
             linewidth = 1) +
  theme_classic()
  
```

### Histogram (Base R)

```{r}
hist(cl.order$LEN_MC)
  abline(v=mean(cl.order$LEN_MC),lwd=3, col = "steelblue")
```

### Density plot (Base R)

```{r}
plot(density(cl.order$LEN_MC))
  abline(v=mean(cl.order$LEN_MC),lwd=3, col = "steelblue")
```
:::

### The median {.smaller}

The `median()` function computes the "the halfway point of the data (50%
of the data are above the median; 50% of the data are below"
[@winter_statistics_2020: 58].

$$
\tilde{x}_{0.5} = 
\begin{cases}
x_{((n+1)/2)} & \text{if } n \text{ is odd.} \\
\frac{1}{2}(x_{n/2}+x_{(n/2+1)}) & \text{if } n \text{ is even.}
\end{cases} 
$$

```{r, echo = TRUE, output = TRUE}
median(cl.order$LEN_MC)
```


The median of `LEN_MC` is represented by the red vertical line.

::: panel-tabset
### Histogram (ggplot2)

```{r, echo = TRUE, output = TRUE}
cl.length.hist +
  # Add mean
  geom_vline(aes(xintercept = mean(LEN_MC)), color = "steelblue", linewidth = 1) +
  # Add median
  geom_vline(aes(xintercept = median(LEN_MC)), color = "red", linewidth = 1) +
  theme_classic()
```

### Density plot (ggplot2)

```{r, echo = TRUE, output = TRUE}
cl.length.dens +
  # Add mean
  geom_vline(aes(xintercept = mean(LEN_MC)), color = "steelblue", linewidth = 1) +
  # Add median
  geom_vline(aes(xintercept = median(LEN_MC)), color = "red", linewidth = 1) +
  theme_classic()
```

### Histogram (Base R)

```{r}
hist(cl.order$LEN_MC)
  abline(v=mean(cl.order$LEN_MC),lwd=3, col = "steelblue")
  abline(v=median(cl.order$LEN_MC),lwd=3, col = "red")
```

### Density plot (Base R)

```{r}
plot(density(cl.order$LEN_MC))
  abline(v=mean(cl.order$LEN_MC),lwd=3, col = "steelblue")
  abline(v=mean(cl.order$LEN_MC),lwd=3, col = "red")
```
:::

### Sample variance and standard deviation {.smaller}

In order to assess how well the mean represents the data, it is
instructive to compute the **variance** `var()` and the **standard
deviation** `sd()` for a sample.

The sample variance $s^2$ is defined as

$$s^2 = \frac{1}{n}\sum_{i=1}^n{(x_i - \bar{x})^2}. $$

In other words, it stands for the average squared deviation of all
observations from the sample mean.

```{r, echo = TRUE, output = TRUE}
var(cl.order$LEN_MC)
```

Correspondingly, the standard deviation of the mean is the square root
of the variance [cf. @heumann_introduction_2022: 51-2].

$$ s = \sqrt{\frac{1}{n}\sum_{i=1}^n{(x_i - \bar{x})^2}} $$

```{r, echo = TRUE, output = TRUE}
sd(cl.order$LEN_MC)
```

**Visualisation**:

::: panel-tabset
#### Example 1

```{r, echo = TRUE, output = TRUE}
cl.length.hist +
  # Add verticle line for the mean
  geom_vline(aes(xintercept = mean(LEN_MC)), color = "steelblue", linewidth = 1) +
  # Add -1sd
  geom_vline(aes(xintercept = mean(LEN_MC) - sd(LEN_MC)), color = "orange", linewidth = 1) +
  # Add +1sd
  geom_vline(aes(xintercept = mean(LEN_MC) + sd(LEN_MC)), color = "orange", linewidth = 1) +
  theme_classic()
```

#### Example 2

```{r, echo = TRUE, output = TRUE}

# Create data frame with mean and sd for each clause ORDER

cl.order %>% 
  # Select variables of interest
  select(ORDER, LEN_MC) %>% 
  # Group results of following operations by ORDER
  group_by(ORDER) %>% 
    # Create grouped summary of mean and sd for each ORDER
    summarise(mean = mean(LEN_MC),
                sd = sd(LEN_MC)) -> cl_mean_sd; cl_mean_sd

# Plot results 

ggplot(cl_mean_sd, aes(x = ORDER, y = mean)) +
  # Barplot with a specific variable mapped onto y-axis
  geom_col() +
  # Add mean and standard deviation to the plot
  geom_errorbar(aes(x = ORDER,
                    ymin = mean-sd,
                    ymax = mean+sd), width = .2) +
  theme_classic() +
  labs(y = "Mean length of main clauses", x = "Clause order")
  
```
:::

### Quantiles

While `median()` divides the data into two equal sets (i.e., two 50%
quantiles), the `quantile()` function makes it possible to partition the
data further.

```{r, echo = TRUE}
quantile(cl.order$LEN_MC)
```

`quantile(x, 0)` and `quantile(x, 1)` thus show the minimum and maximum
values, respectively.

```{r, echo = TRUE}
quantile(cl.order$LEN_MC, 0)
quantile(cl.order$LEN_MC, 1)
```

### Quartiles and boxplots

Consider the distribution of clause length by clause order:

::: panel-tabset
#### Boxplot (Base R)

```{r}
boxplot(LEN_MC ~ ORDER, cl.order)
```

#### Boxplot (ggplot2)

```{r, echo = TRUE, output = TRUE}
ggplot(cl.order, aes(x = ORDER, y = LEN_MC)) +
  geom_boxplot() +
  theme_classic()
```
:::

Compare it to the corresponding rotated density plot:

```{r, echo = TRUE, output = TRUE}

ggplot(cl.order, aes(x = LEN_MC, fill = ORDER)) +
  geom_density(alpha = 0.5) +
  coord_flip() +
  theme_classic()

```

## Visualising mixed data

### A numerical and categorical variable

-   Boxplot with `geom_boxplot()`

```{r, echo = TRUE, output = TRUE}
ggplot(cl.order, aes(x = ORDER, y = LEN_MC)) +
  geom_boxplot()
```

-   Densitiy plot using the optional arguments `color` and/or `fill`

```{r, echo = TRUE, output = TRUE}
ggplot(cl.order, aes(x = LEN_MC, fill = ORDER)) +
  geom_density(alpha = 0.5)
```

-   A barplot with `geom_col()`

```{r, echo = TRUE, output = TRUE}
ggplot(cl.order, aes(x = ORDER, y = LEN_MC)) +
  geom_col(aes(x = ORDER, y = LEN_MC))
```

### Multivariate plots

-   Advanced scatterplot with four variables: `LEN_MC` (x), `LEN_SC`
    (y), `ORDER` (colour) and `SUBORDTYPE` (shape)

```{r, echo = TRUE, output = TRUE}
# 4 variables
ggplot(cl.order, aes(x = LEN_MC, y = LEN_SC)) +
  geom_point(aes(color = ORDER, shape = SUBORDTYPE))
```

-   Facets

```{r, echo = TRUE, output = TRUE}
# 5 variables
ggplot(cl.order, aes(x = LEN_MC, y = LEN_SC)) +
  geom_point(aes(color = ORDER, shape = SUBORDTYPE)) +
  facet_wrap(~MORETHAN2CL)
```
