---
title: "Data types"
author:
  name: "Vladimir Buskin" 
  orcid: "0009-0005-5824-1012"
  affiliation: 
    name: "Catholic University of Eichstätt-Ingolstadt"
    department: "English Language and Linguistics"
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
bibliography: R.bib
---

## Recommended reading

> @heumann_introduction_2022: Chapter 1
>
> @baguleySeriousStatsGuide2012: Chapter 1
>
> @agrestiFoundationsStatisticsData2022: Chapter 1.2

## Preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Please download the file `Paquot_Larsson_2020_data.xlsx`
[@paquotDescriptiveStatisticsVisualization2020][^variables-1] and store
it in your working directory.

[^variables-1]: The supplementary materials can be downloaded from the
    publisher's
    [website](https://link.springer.com/chapter/10.1007/978-3-030-46216-1_17)
    \[Last accessed April 28, 2024\]. Note that the dataset has been
    originally compiled by @gries_statistics_2013.

```{r, echo = TRUE, output = FALSE}
# Load libraries
library(readxl)
library(tidyverse)

# Load data
cl.order <- read_xlsx("Paquot_Larsson_2020_data.xlsx")

# Inspect data
str(cl.order)
head(cl.order)

```

::: {.callout-note collapse="true" title="What's in the file? The `str()` function"}
The easiest way to get a general overview of the full data set is to
apply the `str()` function to the respective data frame.

```{r}
str(cl.order)
```

This shows us that the data frame has 8 columns, as the `$` operators
indicate (`$ Case`, `$ ORDER`, ...). The column names are followed by

-   the data type (`num` for numeric and `chr` for character strings)

-   the number of values (`` `[1:403]` ``) and

-   the first few observations.

Another intuitive way to display the structure of a data matrix is to
simply show the first few rows:

```{r}
head(cl.order)
```
:::

::: {.callout-note collapse="true" title="Further details on the variables"}
-   `ORDER`: Does the subordinate clause come before or after the main
    clause? ('sc-mc' vs. 'mc-sc')

-   `SUBORDTYPE`: Is the subordinate clause temporal or causal? ('temp'
    vs. 'caus')

-   `MORETHAN2CL`: Are there most clauses in the sentence than just one
    subordinate clause and one main clause? ('yes' vs. 'no')

-   `LEN_MC`: How many words does the main clause contain? (ratio-scaled
    continuous variable)

-   `LEN_SC`: How many words does the subordinate clause contain?
    (ratio-scaled continuous variable)

-   `LENGTH_DIFF`: What is the length difference in words between the
    main clause and subordinate clause? (ratio-scaled continuous
    variables)
:::

## The big picture: Populations and samples

In order to investigate one or more (extra-)linguistic features, we
first need to collect relevant **observations**, which typically
correspond to linguistic utterances in a corpus-linguistic context.
Since it is not feasible to examine the entirety of all linguistic
utterances ever produced, i.e., the **population**, we rely on specific
subsets of it: **samples**.

To state this more formally, let $\omega$ ('lower-case Omega') denote a
single observation of interest. The full set of theoretically possible
observations is contained in $\Omega$ ('upper-case Omega'), with
$\Omega = \{\omega_1, \omega_2, \dots, \omega_n \}$. Here $n$ represents
the $n$-th observation and should be a natural number. A sample is then
simply a selection of elements from $\Omega$.

## Variables

The concept of the **variable** is very handy in that it allows us to
quantify different aspects our linguistic observations, such the age or
gender of the speaker, the register of the speech situation, the variety
of English, the length of an utterance, a syntactic construction or a
grammatical category, among many other theoretically possible features.
For each observation, we should be able to assign a specific **value**
to our variables; for instance, a variable `Register` could assume a
value such as `informal`, or `Utterance length` a hypothetical value of
`5` (e.g., 5 words).

In the statistical literature, variables are usually represented by
upper-case letters, such as $X$. For each observation
$\omega \in \Omega$, $X$ has a specific outcome $x$. The set $S$
comprises all possible outcomes of $X$, e.g.,

-   $S_{\text{Register}} = \{\text{informal}, \text{formal}, ...\}$,

-   $S_{\text{Utterance length}} = \{1, 2, 3, \dots \}$,

-   ...

::: callout-note
## Number of (unique) values in R

To count the number of items in a vector, which correspond to the total
number $n$ of attested values of $X$, we can use `length()`:

```{r}
length(cl.order$SUBORDTYPE)
```

In fact, it is equivalent to the number of rows in the full data frame:

```{r}
nrow(cl.order)
```

The function `unique()` shows all unique items (= types) in a vector,
which reflect the possible outcomes in $S$:

```{r}
unique(cl.order$SUBORDTYPE)
```
:::

Information on variables and their values is conventionally arranged in
a **data set**, which is essentially a matrix with columns (= variables)
and rows (= observations). Assuming that $n$ denotes the row number and
$p$ the column number, the general form of a dataset is summarised in
@eq-dataset.

$$
\begin{pmatrix}
    \omega & X_1 & X_2 & \cdots &  X_p \\
    1 & x_{11} & x_{12} & \cdots & x_{1p} \\
    2 & x_{21} & x_{22} & \cdots & x_{2p} \\
    \vdots & \vdots & \vdots &  & \vdots \\
    n & x_{n1} & x_{n2} & \cdots & x_{np}
\end{pmatrix}
$$ {#eq-dataset}

### Subtypes

In general, we distinguish between **discrete** variables, which can
only take a limited set of unique values, and **continuous** variables,
which can take infinitely many values within a specified range.

We can further subdivide discrete and continuous variables into into
nominal, ordinal, interval-scaled and ratio-scaled ones:

::: {.callout-note title="Nominal/categorical" collapse="true"}
These variables comprise a limited number of categories which **cannot
be ordered** in a meaningful way. For instance, it does not matter which
value of `SUBORDTYPE` or `MORETHAN2CL` comes first or last:

```{r, echo = TRUE, output = TRUE}
    unique(cl.order$SUBORDTYPE)
    unique(cl.order$MORETHAN2CL)
```
:::

::: {.callout-note title="Ordinal/categorical" collapse="true"}
Ordinal variables are **ordered**. However, the intervals between their
individual values are not interpretable. Heumann
[-@heumann_introduction_2022: 6] provides a pertinent example:

> \[T\]he satisfaction with a product (unsatisfied–satisfied–very
> satisfied) is an ordinal variable because the values this variable can
> take can be ordered but the differences between
> 'unsatisfied–satisfied' and 'satisfied–very satisfied' cannot be
> compared in a numerical way.
:::

::: {.callout-note title="Interval-scaled/continuous" collapse="true"}
-   In the case of **interval**-scaled variables, the differences
    between the values can be interpreted, but their ratios must be
    treated with caution. A temperature of 4°C is 6 degrees warmer than
    -2°C; however, this does not imply that 4°C is three times warmer
    than -2°C. This is because the temperature scale has no true zero
    point; 0°C simply signifies another point on the scale and not the
    absence of temperature altogether.
:::

::: {.callout-note title="Ratio-scaled/continuous" collapse="true"}
-   **Ratio**-scaled variables allow both a meaningful interpretation of
    the differences between their values and (!) of the ratios between
    them. Within the context of clause length, `LENGTH_DIFF` values such
    as 4 and 8 not only suggest that the latter is four units greater
    than the former but also that their ratio $\frac{8}{4} = 2$ is a
    valid way to describe the relationship between these values. Here a
    `LENGTH_DIFF` of 0 can be clearly viewed as the absence of a length
    difference.
:::

### Dependent vs. independent variables

In empirical studies, it is often of interest whether one variable leads
to changes in the values of another variable. When exploring such
**associations** (or **correlations**), we need to take another
heuristic step to clarify the direction of the influence.

In a linguistic context, we denote the variable whose usage patterns
we'd like to explain as the **dependent** or **response** variable. A
list of possible dependent variables is provided in the section on
[Linguistic variables](Linguistic_variables.qmd)).

Its outcomes are said to depend on one or more **independent
variables**. These are also often referred to as **explanatory**
variables as they are supposed to explain variation in the response
variable. These can be `AGE`, `SEX` or the `VARIETY` of English at hand.

### Exercises

::: {#exr-var-1}
Identify the variable types (nominal, ordinal etc.) for all columns in
the `cl.order` dataset.
:::

::: {#exr-var-2}
Consider the general form of the data matrix from @eq-dataset.

1.  What would the labels $\omega$, $X$ and $x$ correspond to in the
    `cl.order` dataset?

2.  What would be $\Omega$ and $S$?
:::

::: {#exr-var-3}
Consider the following statement:

> This paper examines the influence of clause length on the ordering of
> main and subordinate clauses.

1.  What is the dependent variable?

2.  What is/are the independent variable(s)?

3.  What would it mean if they were reversed?
:::
