---
title: "Statistical variables"
author: Vladimir Buskin
institute: Catholic University of Eichstätt-Ingolstadt
format:
  html:
    self-contained: false
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
bibliography: R.bib
---

## Preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Please download the file "Paquot_Larsson_2020_data.xlsx"
[@paquotDescriptiveStatisticsVisualization2020][^variables-1] and store
it in your working directory.

[^variables-1]: The supplementary materials can be downloaded from the
    publisher's
    [website](https://link.springer.com/chapter/10.1007/978-3-030-46216-1_17)
    \[Last accessed April 28, 2024\].

```{r, echo = TRUE, output = TRUE}
# Libraries
library("nycflights13")
library("readxl")
library("tidyverse")
library("ggthemes")

# Load data
cl.order <- read_xlsx("Paquot_Larsson_2020_data.xlsx")

# Inspect data
str(cl.order)
head(cl.order)

```

## Types of variables

The concept of the **variable** allows us to quantify various aspects of
our observations.

-   **nominal/categorical**: These variables have a limited number of
    levels which cannot be ordered in a meaningful way. For instance, it
    does not matter which value of `SUBORDTYPE` or `MORETHAN2CL` comes
    first or last:

    ```{r, echo = TRUE, output = TRUE}
    unique(cl.order$SUBORDTYPE)
    unique(cl.order$MORETHAN2CL)
    ```

-   **ordinal**: Such variables can be ordered, but the intervals
    between their individuals values are not meaningful. Heumann
    [-@heumann_introduction_2022: 6] provides a pertinent example:

    "\[T\]he satisfaction with a product (unsatisfied–satisfied–very
    satisfied) is an ordinal variable because the values this variable
    can take can be ordered but the differences between
    'unsatisfied–satisfied' and 'satisfied–very satisfied' cannot be
    compared in a numerical way".

-   In the case of **interval**-scaled variables, the differences
    between the values can be interpreted, but their ratios must be
    treated with caution. A temperature of 4°C is 6 degrees warmer than
    -2°C; however, this does not imply that 4°C is three times warmer
    than -2°C. This is because the temperature scale has no true zero
    point; 0°C simply signifies another point on the scale and not the
    absence of temperature altogether.

-   **Ratio**-scaled variables allow both a meaningful interpretation of
    the differences between their values and (!) of the ratios between
    them. Within the context of clause length, `LENGTH_DIFF` values such
    as 4 and 8 not only suggest that the latter is four units greater
    than the former but also that their ratio $\frac{8}{4} = 2$ is a
    valid way to describe the relationship between these values. Here a
    `LENGTH_DIFF` of 0 can be clearly viewed as the absence of a length
    difference.
