---
title: "Data types"
author: Vladimir Buskin
institute: Catholic University of Eichstätt-Ingolstadt
format:
  html:
    self-contained: true
    theme: default
    toc: true
    number-sections: true
    slide-number: true
    incremental: false
    slide-level: 3
    scrollable: true
editor: visual
bibliography: R.bib
---

## Recommended reading

> @heumann_introduction_2022: Chapter 1.2
>
> @agrestiFoundationsStatisticsData2022: Chapter 1.2

## Preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Please download the file `Paquot_Larsson_2020_data.xlsx`
[@paquotDescriptiveStatisticsVisualization2020][^variables-1] and store
it in your working directory.

[^variables-1]: The supplementary materials can be downloaded from the
    publisher's
    [website](https://link.springer.com/chapter/10.1007/978-3-030-46216-1_17)
    \[Last accessed April 28, 2024\]. Note that the dataset has been
    originally compiled by @gries_statistics_2013.

```{r, echo = TRUE, output = FALSE}
# Libraries
library("readxl")
library("tidyverse")

# Load data
cl.order <- read_xlsx("Paquot_Larsson_2020_data.xlsx")

# Inspect data
str(cl.order)
head(cl.order)

```

::: {.callout-note collapse="true" title="What's in the file? The `str()` function"}
The easiest way to get a general overview of the full data set is to
apply the `str()` function to the respective data frame.

```{r}
str(cl.order)
```

This shows us that the data frame has 8 columns, as the `$` operators
indicate (`$ Case`, `$ ORDER`, ...). The column names are followed by

-   the data type (`num` for numeric and `chr` for character strings)

-   the number of values (`` `[1:403]` ``) and

-   the first few observations.

Another intuitive way to display the structure of a data matrix is to
simply show the first few rows:

```{r}
head(cl.order)
```
:::

::: {.callout-note collapse="true" title="Further details on the variables"}
-   `ORDER`: Does the subordinate clause come before or after the main
    clause? ('sc-mc' vs. 'mc-sc')

-   `SUBORDTYPE`: Is the subordinate clause temporal or causal? ('temp'
    vs. 'caus')

-   `MORETHAN2CL`: Are there most clauses in the sentence than just one
    subordinate clause and one main clause? ('yes' vs. 'no')

-   `LEN_MC`: How many words does the main clause contain? (ratio-scaled
    continuous variable)

-   `LEN_SC`: How many words does the subordinate clause contain?
    (ratio-scaled continuous variable)

-   `LENGTH_DIFF`: What is the length difference in words between the
    main clause and subordinate clause? (ratio-scaled continuous
    variables)
:::

## Variables

The concept of the **variable** is very handy in that it allows us to
quantify various aspects their observations. In general, we distinguish
between **discrete** variables, which can only take a limited set of
unique values, and **continuous** variables, which can take infinitely
many values within a specified range.

### Subtypes

We can further subdivide discrete and continuous variables into into
nominal, ordinal, interval-scaled and ratio-scaled ones:

::: {.callout-note title="Nominal/categorical" collapse="true"}
These variables comprise a limited number of categories which **cannot
be ordered** in a meaningful way. For instance, it does not matter which
value of `SUBORDTYPE` or `MORETHAN2CL` comes first or last:

```{r, echo = TRUE, output = TRUE}
    unique(cl.order$SUBORDTYPE)
    unique(cl.order$MORETHAN2CL)
```
:::

::: {.callout-note title="Ordinal/categorical" collapse="true"}
Ordinal variables are **ordered**. However, the intervals between their
individual values are not interpretable. Heumann
[-@heumann_introduction_2022: 6] provides a pertinent example:

> \[T\]he satisfaction with a product (unsatisfied–satisfied–very
> satisfied) is an ordinal variable because the values this variable can
> take can be ordered but the differences between
> 'unsatisfied–satisfied' and 'satisfied–very satisfied' cannot be
> compared in a numerical way.
:::

::: {.callout-note title="Interval-scaled/continuous" collapse="true"}
-   In the case of **interval**-scaled variables, the differences
    between the values can be interpreted, but their ratios must be
    treated with caution. A temperature of 4°C is 6 degrees warmer than
    -2°C; however, this does not imply that 4°C is three times warmer
    than -2°C. This is because the temperature scale has no true zero
    point; 0°C simply signifies another point on the scale and not the
    absence of temperature altogether.
:::

::: {.callout-note title="Ratio-scaled/continuous" collapse="true"}
-   **Ratio**-scaled variables allow both a meaningful interpretation of
    the differences between their values and (!) of the ratios between
    them. Within the context of clause length, `LENGTH_DIFF` values such
    as 4 and 8 not only suggest that the latter is four units greater
    than the former but also that their ratio $\frac{8}{4} = 2$ is a
    valid way to describe the relationship between these values. Here a
    `LENGTH_DIFF` of 0 can be clearly viewed as the absence of a length
    difference.
:::

### Dependent vs. independent variables

In empirical studies, it is often of interest whether one variable leads
to changes in the values of another variable. When exploring such
**associations**, we need to take another heuristic step to clarify the
direction of the influence.

In a linguistic context, we denote the variable whose usage patterns
we'd like to explain as the **dependent** or **response** variable. A
list of possible dependent variables is provided in the section on
[Linguistic variables](Linguistic_variables.qmd)).

Its outcomes are said to depend on one or more **independent
variables**. These are also often referred to as **explanatory**
variables as they are supposed to explain variation in the response
variable. These can be `AGE`, `SEX` or the `VARIETY` of English at hand.

### Exercises

::: {#exr-var-1}
Classify the remaining variables in the `cl.order` dataset.
:::

::: {#exr-var-2}
Consider the following statement:

> This paper examines the influence of clause length on the ordering of
> main and subordinate clauses.

What is the dependent variable? What is/are the independent variable(s)?
:::
